---
title: "endo_rectal_v9_pain_final"
author: "Cameron A. Gorzney, Stephen Johnson, Raunak Vijayakar, Michelle Bland"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries

```{r}
###r version 4.3.2

# # Install specific Bioconductor version
# install.packages("BiocManager")
# 
# BiocManager::install(version = "3.17")
# 
# # Then install packages
# BiocManager::install(c("phyloseq", "microbiome"))
```


```{r}
# packageVersion("")
library(vegan) #remotes::install_version("vegan", version = "2.6-10")
library(ggrepel)
library(readxl) #remotes::install_version("readxl", version = "1.4.5")  
library(phyloseq) #see above
library(ggplot2) #remotes::install_version("ggplot2", version = "4.0.0")  
library(hrbrthemes) #remotes::install_version("hrbrthemes", version = "0.8.7")  
library(ggsci) #remotes::install_version("ggsci", version = "3.2.0")  
library(dplyr) #remotes::install_version("dplyr", version = "1.1.4")  
library(microViz) #see below #needed to install "ggiraph" and "ComplexHeatmap"
# install.packages(
#   "microViz",
#   repos = c(davidbarnett = "https://david-barnett.r-universe.dev", getOption("repos"))
# )
library(microbiome) #see above
library(MicrobiomeStat) #remotes::install_version("MicrobiomeStat", version = "1.0")  
library(GUniFrac) #remotes::install_version("GUniFrac", version = "1.8")
library(picante) #remotes::install_version("picante", version = "1.8.2")
library(ggtext) #remotes::install_version("ggtext", version = "2.6-10")
library(tidyr)
library(purrr)
```

## Process samples- Update paths as needed

```{r}

setwd()

phylo = readRDS("data/Phylo_public.rds")
meta = samdat_tbl(phylo)

#look at metadata
summarize_phyloseq(phylo)

table(meta$Source) #I thought this was a binary variable? We'll use only vaginal and rectal
table(meta$Plate, meta$case_control) #maybe include plate as covariate in models?

```

Add sample library size.

```{r}
depth = data.frame(lib_size = sample_sums(phylo))
depth$SampleID = rownames(depth)
depth$SampleID = sub('.', '', depth$SampleID)

phylo = ps_join(x = phylo, y = depth, by = "SampleID", type = "inner", match_sample_names = NULL)

sample_data(phylo)$subject = as.factor(sample_data(phylo)$subject)
```


```{r}
unique(meta$Source)
unique(meta$Amp.Status)

phylo_r = phylo |> ps_filter(Source == "rectum") 
temp = sample_data(phylo_r)
unique(temp$Subject.ID)

colSums(table(temp$Subject.ID, temp$case_control) != 0) #check sample size

phylo_r = phylo_r |> ps_filter(Amp.Status == "Amp")
meta_r = samdat_tbl(phylo_r)
unique(meta_r$Subject.ID)

colSums(table(meta_r$Subject.ID, meta_r$case_control) != 0) #check sample size

#list the subjects who were lost after filtering for amplification
unique(temp$Subject.ID[!(temp$Subject.ID %in% meta_r$Subject.ID)])

```

Merge and filter.

```{r}
merged_r = merge_samples(phylo_r, "subject", fun = sum)
# We use the sum function here instead of mean to get correct lib_size.
# The other numeric metadata are completely wrecked, as they would be with any function.

# Keep subject IDs from the merged phyloseq object
sample_data(merged_r)$Subject.ID = sample_names(merged_r)

# Add corrected patient demographics metadata
newdata = read_xlsx(
"data/Newdata_endo_pain.xlsx",col_names = T, sheet = 1)

splits = strsplit(newdata$`Study ID`, split = "-")
splits = lapply(splits, head, 2)
newdata$Subject.ID = sapply(splits, paste, collapse = "")


# Keep all clinical variables from the corrected demographics sheet
newdata = newdata |> 
  dplyr::rename(
    case_control_new = `case vs control`,
    paratubal_cysts_new = `Paratubal Cysts`,
    Gravidity_new = Graviditity,
    Age_new = `Age at surgery`,
    parity_status_new= `Parity status`,
    parity_new = Parity,
    BMI_new = BMI,
    stage_new = stage,
    Severity_new = Severity,
    Endosalpingiosis_new = Endosalpingiosis,
    Adenomyosis_new = Adenomyosis,
    Fibroids_new = Fibroids,
    pain_1st_level_new = pain_1st_level,
    abnormal_bleeding_new = Abnormal_bleeding
  ) |>
  dplyr::select(
    Subject.ID,
    case_control_new,
    BMI_new,
    Age_new,
    Gravidity_new,
    stage_new,
    Severity_new,
    parity_status_new,
    Endosalpingiosis_new,
    Adenomyosis_new,
    Fibroids_new,
    paratubal_cysts_new,
    parity_new,
    pain_1st_level_new,
    abnormal_bleeding_new,
    Race,
    Ethnicity
  ) |>
  dplyr::distinct(Subject.ID, .keep_all = TRUE)

merged_r = ps_join(merged_r, newdata, by = "Subject.ID", type = "left")

meta_pain <- samdat_tbl(merged_r)

pain_subject <- meta_pain %>%
  dplyr::select(Subject.ID, pain_1st_level_new) %>%
  dplyr::distinct() %>%
  dplyr::mutate(
    pain_score = as.numeric(pain_1st_level_new),
    pain_score = dplyr::case_when(
      is.na(pain_score) ~ NA_real_,
      pain_score == 99 ~ NA_real_,
      pain_score < 0 ~ NA_real_,
      pain_score > 10 ~ NA_real_,
      TRUE ~ pain_score
    ),
    pain_severity = dplyr::case_when(
      pain_score == 0 ~ "No pain",
      pain_score >= 1 & pain_score <= 3 ~ "Mild pain",
      pain_score >= 4 & pain_score <= 6 ~ "Moderate pain",
      pain_score >= 7 & pain_score <= 10 ~ "Severe pain",
      TRUE ~ NA_character_
    ),
    pain_severity = factor(
      pain_severity,
      levels = c("No pain", "Mild pain", "Moderate pain", "Severe pain")
    ),
    pain_binary = dplyr::case_when(
      pain_score <= 3 ~ "No/Mild",
      pain_score >= 4 ~ "Moderate/Severe",
      TRUE ~ NA_character_
    ),
    pain_binary = factor(
      pain_binary,
      levels = c("No/Mild", "Moderate/Severe")
    )
  ) %>%
  dplyr::filter(!is.na(pain_score)) %>%
  dplyr::select(Subject.ID, pain_score, pain_severity, pain_binary)

table(pain_subject$pain_score)
table(pain_subject$pain_severity)
table(pain_subject$pain_binary)


merged_r = ps_join(merged_r, pain_subject, by = "Subject.ID", type = "inner")

# Number of unique participants with BOTH microbiome + pain data
n_distinct(sample_data(merged_r)$Subject.ID)

# Check that corrected demographics are still present after join
merged_meta_check = samdat_tbl(merged_r)

colnames(merged_meta_check)

table(merged_meta_check$case_control_new, useNA = "ifany")
summary(merged_meta_check$BMI_new)
summary(merged_meta_check$Age_new)
summary(merged_meta_check$parity_new)
table(merged_meta_check$Severity_new, useNA = "ifany")


merged_r <- merged_r %>%
  ps_filter(lib_size > 2000) %>%
  tax_filter(min_prevalence = 2)

merged_meta_r = samdat_tbl(merged_r)

merged_meta_r$case_control_new = as.factor(merged_meta_r$case_control_new)
merged_meta_r$case_control_new = relevel(merged_meta_r$case_control_new, ref = "control")
merged_meta_r$Fibroids_new = as.factor(merged_meta_r$Fibroids_new)
merged_meta_r$paratubal_cysts_new = as.factor(merged_meta_r$paratubal_cysts_new)
merged_meta_r$Endosalpingiosis_new = as.factor(merged_meta_r$Endosalpingiosis_new)
merged_meta_r$Adenomyosis_new = as.factor(merged_meta_r$Adenomyosis_new)


#check for library size confounding
summary(merged_meta_r$lib_size)
hist(merged_meta_r$lib_size)
boxplot(merged_meta_r$lib_size ~ merged_meta_r$case_control_new, notch = F)
wilcox.test(merged_meta_r$lib_size ~ merged_meta_r$case_control_new)
boxplot(merged_meta_r$lib_size ~ merged_meta_r$pain_severity, notch = F)
kruskal.test(merged_meta_r$lib_size ~ merged_meta_r$pain_severity)


#rarefy
set.seed(123)
rare_r = rarefy_even_depth(merged_r, replace = F) #rarefy to min depth or 2k, whichever greater

```
### Check matched samples
```{r}
# Rectal subjects after Amp + merge + pain join + lib_size/tax filtering
rectal_pain_subjects <- samdat_tbl(merged_r) %>%
  dplyr::filter(!is.na(pain_severity)) %>%
  dplyr::distinct(Subject.ID) %>%
  dplyr::pull(Subject.ID)


phylo_v <- phylo |> 
  ps_filter(Source == "vagina") |>
  ps_filter(Amp.Status == "Amp")

merged_v <- merge_samples(phylo_v, "subject", fun = sum)

sample_data(merged_v)$Subject.ID <- sample_names(merged_v)

merged_v <- ps_join(merged_v, newdata, by = "Subject.ID", type = "left")

meta_pain_v <- samdat_tbl(merged_v)

pain_subject_v <- meta_pain_v %>%
  dplyr::select(Subject.ID, pain_1st_level_new) %>%
  dplyr::distinct() %>%
  dplyr::mutate(
    pain_score = as.numeric(pain_1st_level_new),
    pain_score = dplyr::case_when(
      is.na(pain_score) ~ NA_real_,
      pain_score == 99 ~ NA_real_,
      pain_score < 0 ~ NA_real_,
      pain_score > 10 ~ NA_real_,
      TRUE ~ pain_score
    ),
    pain_severity = dplyr::case_when(
      pain_score == 0 ~ "No pain",
      pain_score >= 1 & pain_score <= 3 ~ "Mild pain",
      pain_score >= 4 & pain_score <= 6 ~ "Moderate pain",
      pain_score >= 7 & pain_score <= 10 ~ "Severe pain",
      TRUE ~ NA_character_
    ),
    pain_severity = factor(
      pain_severity,
      levels = c("No pain", "Mild pain", "Moderate pain", "Severe pain")
    )
  ) %>%
  dplyr::filter(!is.na(pain_score)) %>%
  dplyr::select(Subject.ID, pain_score, pain_severity)

merged_v <- ps_join(merged_v, pain_subject_v, by = "Subject.ID", type = "inner")

merged_v <- merged_v %>%
  ps_filter(lib_size > 2000) %>%
  tax_filter(min_prevalence = 2)

merged_meta_v <- samdat_tbl(merged_v)


# Vaginal subjects after Amp + merge + pain join + lib_size/tax filtering
vaginal_pain_subjects <- samdat_tbl(merged_v) %>%
  dplyr::filter(!is.na(pain_severity)) %>%
  dplyr::distinct(Subject.ID) %>%
  dplyr::pull(Subject.ID)

# Subjects with BOTH rectal and vaginal swabs in pain cohort
matched_pain_subjects <- intersect(rectal_pain_subjects, vaginal_pain_subjects)

length(matched_pain_subjects)
matched_pain_subjects


matched_pain_summary <- samdat_tbl(merged_r) %>%
  dplyr::filter(Subject.ID %in% matched_pain_subjects) %>%
  dplyr::distinct(Subject.ID, pain_severity, case_control_new) %>%
  dplyr::arrange(pain_severity, Subject.ID)

table(matched_pain_summary$pain_severity)
table(matched_pain_summary$case_control_new, matched_pain_summary$pain_severity)


# Subjects with BOTH rectal and vaginal data
matched_subjects <- intersect(rectal_pain_subjects, vaginal_pain_subjects)

# Total matched participants
length(matched_subjects)
```

### Checking other diagnoses against Pain Severity
```{r}
categorical_pvals <- tibble::tibble(
  Variable = c(
    "Adenomyosis",
    "Endosalpingiosis",
    "Fibroids",
    "Endometriosis",
    "Paratubal Cysts",
    "Abnormal Bleeding",
    "Race",
    "Ethnicity"
    
  ),
  p_value = c(
    fisher.test(table(merged_meta_r$Adenomyosis_new,
                      merged_meta_r$pain_severity))$p.value,

    fisher.test(table(merged_meta_r$Endosalpingiosis_new,
                      merged_meta_r$pain_severity))$p.value,

    fisher.test(table(merged_meta_r$Fibroids_new,
                      merged_meta_r$pain_severity))$p.value,
    
    fisher.test(table(merged_meta_r$case_control_new,
                      merged_meta_r$pain_severity))$p.value,

    fisher.test(table(merged_meta_r$paratubal_cysts_new,
                      merged_meta_r$pain_severity))$p.value,

    fisher.test(table(merged_meta_r$abnormal_bleeding_new,
                      merged_meta_r$pain_severity))$p.value,

    fisher.test(table(merged_meta_r$Race,
                      merged_meta_r$pain_severity))$p.value,

    fisher.test(table(merged_meta_r$Ethnicity,
                      merged_meta_r$pain_severity))$p.value
  )
) %>%
  dplyr::mutate(
    FDR = p.adjust(p_value, method = "BH")
  
)

categorical_pvals

### Continous Summary
continuous_summary <- merged_meta_r %>%
  dplyr::filter(
    !is.na(pain_severity),
    pain_severity %in% c("No pain", "Mild pain", "Moderate pain", "Severe pain")
  ) %>%
  dplyr::select(
    Severity_new,
    Age_new,
    BMI_new,
    Gravidity_new,
    parity_new
  ) %>%
  tidyr::pivot_longer(
    cols = c(Age_new, BMI_new, Gravidity_new, parity_new),
    names_to = "Variable",
    values_to = "Value"
  ) %>%
  dplyr::group_by(Variable, Severity_new) %>%
  dplyr::summarise(
    mean = mean(Value, na.rm = TRUE),
    sd = sd(Value, na.rm = TRUE),
    mean_sd = paste0(round(mean, 2), " ± ", round(sd, 2)),
    n = sum(!is.na(Value)),
    .groups = "drop"
  )

continuous_summary

continuous_pvals_pain <- merged_meta_r %>%
  dplyr::filter(!is.na(pain_severity)) %>%
  dplyr::select(
    pain_severity,
    Age_new,
    BMI_new,
    Gravidity_new,
    parity_new
  ) %>%
  tidyr::pivot_longer(
    cols = c(Age_new, BMI_new, Gravidity_new, parity_new),
    names_to = "Variable",
    values_to = "Value"
  ) %>%
  dplyr::group_by(Variable) %>%
  dplyr::summarise(
    p_value = kruskal.test(Value ~ pain_severity)$p.value,
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    FDR = p.adjust(p_value, method = "BH")
  )

continuous_pvals_pain

```

### Alpha Diversity for pain severity

```{r}
set.seed(123)

metric_labels <- c(
  Observed = "Observed ASVs",
  Chao1 = "Chao1 Richness",
  Shannon = "Shannon Diversity Index",
  InvSimpson = "Inverse Simpson Index"
)

tax <- as.data.frame(tax_table(rare_r), stringsAsFactors = FALSE)

plot_richness(rare_r, x="pain_severity", measures=c("Observed", "Chao1", "Shannon", "InvSimpson")) + geom_boxplot()

rich_r <- estimate_richness(
  rare_r,
  measures = c("Observed", "Chao1", "Shannon", "InvSimpson")
)

rich_r$pain_severity <- sample_data(rare_r)$pain_severity

rich_r <- rich_r %>%
  mutate(
    Subject.ID = sample_data(rare_r)$Subject.ID,
    pain_score = sample_data(rare_r)$pain_score,
    pain_severity = factor(
      sample_data(rare_r)$pain_severity,
      levels = c("No pain", "Mild pain", "Moderate pain", "Severe pain"),
      labels = c("No pain\n(n = 6)", "Mild pain\n(n = 8)", "Moderate pain\n(n = 7)", "Severe pain\n(n = 9)")
    ),
    case_control = sample_data(rare_r)$case_control_new
  )

table(rich_r$pain_severity)

alpha_results <- tibble(
  Metric = c("Observed", "Chao1", "Shannon", "InvSimpson"),
  Pvalue = c(
    kruskal.test(Observed ~ pain_severity, data = rich_r)$p.value,
    kruskal.test(Chao1 ~ pain_severity, data = rich_r)$p.value,
    kruskal.test(Shannon ~ pain_severity, data = rich_r)$p.value,
    kruskal.test(InvSimpson ~ pain_severity, data = rich_r)$p.value
  )
) %>%
  mutate(
    FDR = p.adjust(Pvalue, method = "fdr")#,
    #p_label = paste0("Kruskal-Wallis p = ", signif(Pvalue, 3))
  )

alpha_results

plot_alpha_metric <- function(metric_name) {
  
  p_value <- alpha_results %>%
    filter(Metric == metric_name) %>%
    pull(Pvalue)
  
  p_lab <- bquote("Kruskal-Wallis " * italic(p) == .(signif(p_value, 3)))
  
  ggplot(
    rich_r,
    aes(
      x = pain_severity,
      y = .data[[metric_name]],
      fill = pain_severity
    )
  ) +
    geom_boxplot(
      outlier.shape = NA,
      alpha = 0.75,
      color = "black"
    ) +
    geom_jitter(
      width = 0.12,
      size = 3,
      alpha = 0.85,
      color = "black"
    ) +
    scale_fill_manual(
      values = c(
        "No pain\n(n = 6)" = "grey",
        "Mild pain\n(n = 8)" = "#FDBB84",
        "Moderate pain\n(n = 7)" = "#EF6548",
        "Severe pain\n(n = 9)" = "#990000"
      )
    ) +
    theme_classic(base_size = 14) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(angle = 45, hjust = 1, size = 18, face = "bold"),
      axis.text.y = element_text(size = 18),
      axis.title.x = element_text(size = 18),
      axis.title.y = element_text(size = 18, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 24, face = "bold"),
      panel.background = element_rect(fill = "transparent", colour = NA),
      plot.background = element_rect(fill = "transparent", colour = NA)
    ) +
    labs(
      x = "Pain Severity",
      y = metric_labels[[metric_name]],
      #title = paste0(metric_labels[[metric_name]], " by pain severity"),
      subtitle = p_lab
    )
}
p_observed <- plot_alpha_metric("Observed")
p_chao1 <- plot_alpha_metric("Chao1")
p_shannon <- plot_alpha_metric("Shannon")
p_invsimpson <- plot_alpha_metric("InvSimpson")

p_observed
p_chao1
p_shannon
p_invsimpson
```

```{r}
# Save figures

ggsave(
  filename = "Observed_alpha_diversity.png",
  plot = p_observed,
  path = fig_dir,
  width = 6,
  height = 5.5,
  units = "in",
  dpi = 600,
  bg = "transparent"
)

ggsave(
  filename = "Observed_alpha_diversity.pdf",
  plot = p_observed,
  path = fig_dir,
  width = 6,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)

ggsave(
  filename = "Chao1_alpha_diversity.png",
  plot = p_chao1,
  path = fig_dir,
  width = 6,
  height = 5.5,
  units = "in",
  dpi = 600,
  bg = "transparent"
)

ggsave(
  filename = "Chao1_alpha_diversity.pdf",
  plot = p_chao1,
  path = fig_dir,
  width = 6,
  height = 5.5,
  units = "in",
  dpi = 600,
  bg = "transparent"
)

ggsave(
  filename = "Shannon_alpha_diversity.png",
  plot = p_shannon,
  path = fig_dir,
  width = 6,
  height = 5.5,
  units = "in",
  dpi = 600,
  bg = "transparent"
)

ggsave(
  filename = "Shannon_alpha_diversity.pdf",
  plot = p_shannon,
  path = fig_dir,
  width = 6,
  height = 5.5,
  units = "in",
  dpi = 600,
  bg = "transparent"
)
ggsave(
  filename = "InvSimpson_alpha_diversity.png",
  plot = p_invsimpson,
  path = fig_dir,
  width = 6,
  height = 5.5,
  units = "in",
  dpi = 600,
  bg = "transparent"
)

ggsave(
  filename = "InvSimpson_alpha_diversity.pdf",
  plot = p_invsimpson,
  path = fig_dir,
  width = 6,
  height = 5.5,
  units = "in",
  dpi = 600,
  bg = "transparent"
)
```


Phylogenetic diversity from `picante`.

```{r}
set.seed(123)

pd_result <- pd(
  otu_table(rare_r),
  phy_tree(rare_r),
  include.root = TRUE
)

pd_df <- pd_result %>%
  as.data.frame()

pd_df$Subject.ID <- rownames(pd_df)

pd_df$pain_severity <- sample_data(rare_r)$pain_severity

pd_df <- pd_df %>%
  mutate(
    pain_severity = factor(
      pain_severity,
      levels = c("No pain", "Mild pain", "Moderate pain", "Severe pain"),
      labels = c("No pain\n(n = 6)", "Mild pain\n(n = 8)", "Moderate pain\n(n = 7)", "Severe pain\n(n = 9)")
    )
  )

pd_test <- kruskal.test(
  PD ~ pain_severity,
  data = pd_df
)

pd_test

pd_label <- bquote("Kruskal-Wallis " * italic(p) == .(signif(pd_test$p.value, 3)))

pd_plot <- ggplot(
  pd_df,
  aes(
    x = pain_severity,
    y = PD,
    fill = pain_severity
  )
) +
  geom_boxplot(
    outlier.shape = NA,
    alpha = 0.75,
    color = "black"
  ) +
  geom_jitter(
    width = 0.12,
    size = 3,
    alpha = 0.85,
    color = "black"
  ) +
  scale_fill_manual(
    values = c(
      "No pain\n(n = 6)" = "grey",
      "Mild pain\n(n = 8)" = "#FDBB84",
      "Moderate pain\n(n = 7)" = "#EF6548",
      "Severe pain\n(n = 9)" = "#990000"
    )
  ) +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 18, face = "bold"),
    axis.text.y = element_text(size = 18),
      axis.title.x = element_text(size = 18),
      axis.title.y = element_text(size = 18, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 24, face = "bold"),
      panel.background = element_rect(fill = "transparent", colour = NA),
      plot.background = element_rect(fill = "transparent", colour = NA)
    ) +
  labs(
    x = "Pain Severity",
    y = "Faith's Phylogenetic Diversity",
    #title = "Phylogenetic diversity by pain severity",
    subtitle = pd_label
  )
pd_plot
```

# Save Figures
```{r}
ggsave(
  filename = "Faith_PD.png",
  plot = pd_plot,
  path = fig_dir,
  width = 6,
  height = 5.5,
  units = "in",
  dpi = 600,
  bg = "transparent"
)

ggsave(
  filename = "Faith_PD.pdf",
  plot = pd_plot,
  path = fig_dir,
  width = 6,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)
```

### New beta diversity for pain severity, with plots

```{r}
set.seed(123)

# Match metadata to rare_r sample order
merged_meta_r <- merged_meta_r %>%
  filter(Subject.ID %in% sample_names(rare_r)) %>%
  arrange(match(Subject.ID, sample_names(rare_r)))

merged_meta_r$pain_severity <- sample_data(rare_r)[merged_meta_r$Subject.ID, ]$pain_severity

merged_meta_r <- merged_meta_r %>%
  mutate(
    pain_severity = factor(
      pain_severity,
      levels = c("No pain", "Mild pain", "Moderate pain", "Severe pain"),
      labels = c(
        "No pain (n = 6)",
        "Mild pain (n = 8)",
        "Moderate pain (n = 7)",
        "Severe pain (n = 9)"
      )
    )
  )

table(merged_meta_r$pain_severity, useNA = "ifany")

stopifnot(all(merged_meta_r$Subject.ID == sample_names(rare_r)))

# ----------------------------------------------------------
# Build distance matrices
# ----------------------------------------------------------

tree_r <- phy_tree(rare_r)

otu_r <- as(otu_table(rare_r), "matrix")

if (taxa_are_rows(rare_r)) {
  otu_r <- t(otu_r)
}

otu_r <- otu_r[merged_meta_r$Subject.ID, ]

unifracs <- GUniFrac(
  otu.tab = otu_r,
  tree = tree_r,
  alpha = c(0, 0.5, 1)
)$unifracs

uni_dist <- as.dist(unifracs[, , "d_UW"])
guni_dist <- as.dist(unifracs[, , "d_0.5"])
wuni_dist <- as.dist(unifracs[, , "d_1"])
jac_dist <- vegdist(otu_r, method = "jaccard", binary = TRUE)
bray_dist <- vegdist(otu_r, method = "bray")

### Prepare pseudocounts for aitchison
# Genus-level non-rarefied counts
merged_r_genus <- tax_glom(merged_r, taxrank = "Genus")

genus_counts <- as.data.frame(otu_table(merged_r_genus))

# Make sure rows = subjects and columns = taxa
if (taxa_are_rows(merged_r_genus)) {
  genus_counts <- as.data.frame(t(genus_counts))
}

# Add clean genus names as column names
genera <- as.data.frame(tax_table(merged_r_genus))
genus_names <- as.character(genera$Genus)
genus_names[is.na(genus_names) | genus_names == ""] <- "unclassified"
genus_names <- make.unique(genus_names)

colnames(genus_counts) <- genus_names

# CLR transform
genus_counts_pseudo <- genus_counts + 1
clr_mat <- compositions::clr(genus_counts_pseudo)

# Aitchison PCA
pca1 <- prcomp(clr_mat, center = TRUE, scale. = FALSE)
summary(pca1)

scores <- as.data.frame(pca1$x)
scores$Subject.ID <- rownames(scores)

# Aitchison distance
aitch <- vegan::vegdist(clr_mat, method = "euclidean")
# ----------------------------------------------------------
# PERMANOVA
# ----------------------------------------------------------

uni_test <- adonis3(uni_dist ~ pain_severity, data = merged_meta_r)
guni_test <- adonis3(guni_dist ~ pain_severity, data = merged_meta_r)
jac_test <- adonis3(jac_dist ~ pain_severity, data = merged_meta_r)
wuni_test <- adonis3(wuni_dist ~ pain_severity, data = merged_meta_r)
bray_test <- adonis3(bray_dist ~ pain_severity, data = merged_meta_r)
aitch_test <- adonis3(aitch ~ pain_severity, data = merged_meta_r)

beta_results <- data.frame(
  Metric = c(
    "UniFrac",
    "Generalized UniFrac",
    "Jaccard",
    "Weighted UniFrac",
    "Bray-Curtis",
    "Aitchison Distance"
  ),
  R2 = c(
    uni_test$aov.tab$R2[1],
    guni_test$aov.tab$R2[1],
    jac_test$aov.tab$R2[1],
    wuni_test$aov.tab$R2[1],
    bray_test$aov.tab$R2[1],
    aitch_test$aov.tab$R2[1]
  ),
  Pvalue = c(
    uni_test$aov.tab$`Pr(>F)`[1],
    guni_test$aov.tab$`Pr(>F)`[1],
    jac_test$aov.tab$`Pr(>F)`[1],
    wuni_test$aov.tab$`Pr(>F)`[1],
    bray_test$aov.tab$`Pr(>F)`[1],
    aitch_test$aov.tab$`Pr(>F)`[1]
  )
)

beta_results$FDR <- p.adjust(beta_results$Pvalue, method = "fdr")

beta_results

# ----------------------------------------------------------
# Plot settings
# ----------------------------------------------------------

pain_cols <- c(
  "No pain (n = 6)" = "grey",
  "Mild pain (n = 8)" = "#FDBB84",
  "Moderate pain (n = 7)" = "#EF6548",
  "Severe pain (n = 9)" = "#990000"
)

get_p <- function(metric_name) {
  signif(beta_results$Pvalue[beta_results$Metric == metric_name], 3)
}

get_adj_p <- function(metric_name) {
  signif(beta_results$FDR[beta_results$Metric == metric_name], 3)
}

# ----------------------------------------------------------
# PCoA plots from phyloseq-supported distances
# ----------------------------------------------------------
sample_data(rare_r)$pain_severity <- factor(
  sample_data(rare_r)$pain_severity,
  levels = c("No pain", "Mild pain", "Moderate pain", "Severe pain"),
  labels = c(
    "No pain (n = 6)",
    "Mild pain (n = 8)",
    "Moderate pain (n = 7)",
    "Severe pain (n = 9)"
  )
)

ord_uni <- ordinate(rare_r, "PCoA", "unifrac")

p_uni <- plot_ordination(rare_r, ord_uni, color = "pain_severity") +
  stat_ellipse(
    aes(group = pain_severity),
    linewidth = 0.5,
    show.legend = FALSE
  ) +
  geom_point(size = 3.5) +
  scale_color_manual(values = pain_cols) +
  theme_classic(base_size = 16) +
 theme(
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 20, face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 24),
    plot.subtitle = element_text(size = 22, hjust = 0.5),
    legend.title = element_blank(),
    legend.text = element_text(size = 18),
    legend.margin = margin(0, 0, 0, 0),
    
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank()
  ) +
  labs(
    #title = "PCoA using UniFrac",
    x = paste0("PCoA1 [", round(ord_uni$values$Relative_eig[1] * 100, 1), "%]"),
    y = paste0("PCoA2 [", round(ord_uni$values$Relative_eig[2] * 100, 1), "%]"),
    title = bquote("UniFrac"),
    subtitle = bquote("Adjusted " * italic(p) == .(get_adj_p("UniFrac"))),
    color = "Pain Severity"
  )

ord_jac <- ordinate(rare_r, "PCoA", "jaccard")

p_jac <- plot_ordination(rare_r, ord_jac, color = "pain_severity") +
  stat_ellipse(
    aes(group = pain_severity),
    linewidth = 0.5,
    show.legend = FALSE
  ) +
  geom_point(size = 3.5) +
  scale_color_manual(values = pain_cols) +
  theme_classic(base_size = 16) +
  theme(
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 20, face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 24),
    plot.subtitle = element_text(size = 22, hjust = 0.5),
    legend.title = element_blank(),
    legend.text = element_text(size = 18),
    legend.margin = margin(0, 0, 0, 0),
    
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank()
  ) +
  labs(
    #title = "PCoA using Jaccard",
    x = paste0("PCoA1 [", round(ord_jac$values$Relative_eig[1] * 100, 1), "%]"),
    y = paste0("PCoA2 [", round(ord_jac$values$Relative_eig[2] * 100, 1), "%]"),
    title = bquote("Jaccard"),
    subtitle = bquote("Adjusted " * italic(p) == .(get_adj_p("Jaccard"))),
    color = "Pain Severity"
  )

ord_wuni <- ordinate(rare_r, "PCoA", "wunifrac")

p_wuni <- plot_ordination(rare_r, ord_wuni, color = "pain_severity") +
  stat_ellipse(
    aes(group = pain_severity),
    linewidth = 0.5,
    show.legend = FALSE
  ) +
  geom_point(size = 3.5) +
  scale_color_manual(values = pain_cols) +
  theme_classic(base_size = 16) +
  theme(
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 20, face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 24),
    plot.subtitle = element_text(size = 22, hjust = 0.5),
    legend.title = element_blank(),
    legend.text = element_text(size = 18),
    legend.margin = margin(0, 0, 0, 0),
    
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank()
  ) +
  labs(
    #title = "PCoA using Weighted UniFrac",
    x = paste0("PCoA1 [", round(ord_wuni$values$Relative_eig[1] * 100, 1), "%]"),
    y = paste0("PCoA2 [", round(ord_wuni$values$Relative_eig[2] * 100, 1), "%]"),
    title = bquote("Weighted UniFrac"),
    subtitle = bquote("Adjusted " * italic(p) == .(get_adj_p("Weighted UniFrac"))),
    color = "Pain Severity"
  )

ord_bray <- ordinate(rare_r, "PCoA", "bray")

p_bray <- plot_ordination(rare_r, ord_bray, color = "pain_severity") +
  stat_ellipse(
    aes(group = pain_severity),
    linewidth = 0.5,
    show.legend = FALSE
  ) +
  geom_point(size = 3.5) +
  scale_color_manual(values = pain_cols) +
  theme_classic(base_size = 16) +
  theme(
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 20, face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 24),
    plot.subtitle = element_text(size = 22, hjust = 0.5),
    legend.title = element_blank(),
    legend.text = element_text(size = 18),
    legend.margin = margin(0, 0, 0, 0),
    
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank()
  ) +
  labs(
    #title = "PCoA using Bray-Curtis",
    x = paste0("PCoA1 [", round(ord_bray$values$Relative_eig[1] * 100, 1), "%]"),
    y = paste0("PCoA2 [", round(ord_bray$values$Relative_eig[2] * 100, 1), "%]"),
    title = bquote ("Bray-Curtis"),
    subtitle = bquote("Adjusted " * italic(p) == .(get_adj_p("Bray-Curtis"))),
    color = "Pain Severity"
  )


# PCA scores
scores <- as.data.frame(pca1$x)
scores$Subject.ID <- rownames(scores)

# Verify IDs match metadata
setequal(scores$Subject.ID, merged_meta_r$Subject.ID)


# Add pain severity to metadata
scores <- scores %>%
  dplyr::left_join(
    merged_meta_r %>%
      dplyr::select(Subject.ID, pain_severity),
    by = "Subject.ID"
  )

table(scores$pain_severity, useNA = "always")


# Axis labels
var_exp <- (pca1$sdev^2) / sum(pca1$sdev^2)

xlab <- paste0("PC1 [", round(100 * var_exp[1], 1), "%]")
ylab <- paste0("PC2 [", round(100 * var_exp[2], 1), "%]")

#Add Aitchison axes and clusters directly to merged_meta_r
merged_meta_r <- merged_meta_r %>%
dplyr::select(
     -any_of(c(
       "PC1",
       "PC2"
     ))
   ) %>%
   dplyr::left_join(
     scores %>%
       dplyr::select(
         Subject.ID,
         PC1,
         PC2,
       ),
     by = "Subject.ID"
   )

# Aitchison PCA plot
p_aitch <- ggplot(
   merged_meta_r,
   aes(
     x = PC1,
     y = PC2,
     color = pain_severity
   )
 ) +
 
   stat_ellipse(
    aes(group = pain_severity),
    linewidth = 0.5,
    show.legend = FALSE
  ) +

  geom_point(size = 3.5) +
  scale_color_manual(values = pain_cols) +
  theme_classic(base_size = 16) +
  theme(
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 20, face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 24),
    plot.subtitle = element_text(size = 22, hjust = 0.5),
    legend.title = element_blank(),
    legend.text = element_text(size = 18),
    legend.margin = margin(0, 0, 0, 0),

    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),

    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank()
  ) +
  labs(
    x = xlab,
    y = ylab,
    title = bquote("Aitchison Distance"),
    subtitle = bquote("Adjusted " * italic(p) == .(get_adj_p("Aitchison Distance")))
  )


p_aitch

# ----------------------------------------------------------
# Generalized UniFrac PCoA manually from GUniFrac distance
# ----------------------------------------------------------

ord_guni <- cmdscale(guni_dist, eig = TRUE, k = 2)

guni_var = round(ord_guni$eig/sum(ord_guni$eig)*100, 1)

guni_df <- data.frame(
  Subject.ID = rownames(ord_guni$points),
  Axis1 = ord_guni$points[, 1],
  Axis2 = ord_guni$points[, 2]
) %>%
  left_join(
    merged_meta_r %>%
      dplyr::select(Subject.ID, pain_severity),
    by = "Subject.ID"
  )

p_guni <- ggplot(
  guni_df,
  aes(
    x = Axis1,
    y = Axis2,
    color = pain_severity
  )
) +
  stat_ellipse(
    aes(group = pain_severity),
    linewidth = 0.5,
    show.legend = FALSE
  ) +
  geom_point(size = 3.5) +
  scale_color_manual(values = pain_cols) +
  theme_classic(base_size = 16) +
  theme(
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 20, face = "bold"),
    #plot.title = element_text(size = 22, hjust = 0.5),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 24),
    plot.subtitle = element_text(size = 22, hjust = 0.5),
    legend.title = element_blank(),
    legend.text = element_text(size = 18),
    legend.margin = margin(0, 0, 0, 0),
    
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank()
  ) +
  labs(
    #title = "PCoA using Generalized UniFrac",
    x = paste0("PCoA1 [", guni_var[1], "%]"),
    y = paste0("PCoA2 [", guni_var[2], "%]"),
    title = bquote("Generalized UniFrac"),
    subtitle = bquote("Adjusted " * italic(p) == .(get_adj_p("Generalized UniFrac"))),
    color = "Pain Severity"
  )

# Print individual plots (raw p-values)
p_uni
p_guni #not sure why guni didn't have %var on axes. -fixed it above.
p_jac
p_wuni
p_bray
p_aitch
```

### Save beta-diversity plots
```{r}
ggsave(
  filename = "Bray-Curtis.pdf",
  plot = p_bray,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)
ggsave(
  filename = "Bray-Curtis.png",
  plot = p_bray,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  bg = "transparent"
)

ggsave(
  filename = "UniFrac.pdf",
  plot = p_uni,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)

ggsave(
  filename = "UniFrac.png",
  plot = p_uni,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  bg = "transparent"
)

ggsave(
  filename = "WUniFrac.pdf",
  plot = p_wuni,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)
ggsave(
  filename = "WUniFrac.png",
  plot = p_wuni,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  bg = "transparent"
)

ggsave(
  filename = "GUniFrac.pdf",
  plot = p_guni,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)

ggsave(
  filename = "GUniFrac.png",
  plot = p_guni,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  bg = "transparent"
)

ggsave(
  filename = "Jaccard.pdf",
  plot = p_jac,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)

ggsave(
  filename = "Jaccard.png",
  plot = p_jac,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  bg = "transparent"
)

ggsave(
  filename = "Aitch.pdf",
  plot = p_aitch,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)

```

#### PAM clustering

##### Bray-Curtis for pain severity

Calculate BC dissimilarity on rarefied data, then cluster with PAM, k selected using mean silhouette width.
```{r}
#determine number of clusters
library(cluster) #remotes::install_version("cluster", version = "2.1.8.1")
library(scales) #remotes::install_version("scales", version = "1.4.0")

#use sil to choose number of clusters
set.seed(123)
k.max <- 10

SIL = sapply(2:k.max, function(i) {
  results.coef.pam =  pam(bray_dist, i, nstart=50, diss = T)
  mean(silhouette(results.coef.pam$clustering, dist = bray_dist)[,"sil_width"]) })

plot(c(0,SIL), type="b", main = "Selecting k PAM clusters", ylab = "Mean Silhouette Width", xlab = "Clusters")


```

Visualize clusters and pain severity status on PCoA. 

```{r}
set.seed(123)
pam_bc = pam(bray_dist, 2, nstart = 50, diss = T)

# ord = cmdscale(bray_dist, 4, eig = TRUE)
# plot(ord$points, col = clusters, pch = 19, xlab="PCoA1", ylab="PCoA2")

sample_data(rare_r)$Cluster = as.factor(pam_bc$clustering) #add clustering information to ps object's sample data

ord = ordinate(rare_r, "PCoA", "bray")
plot_ordination(rare_r, ord, color = "pain_severity", shape = "Cluster") +
  geom_point(size = 3) + 
  labs(title = "PCoA using Bray-Curtis") +
  theme_ipsum() +
  scale_color_bmj()
```

Correlate rarefied abundances with PCoA axes to look for gradients.
```{r}
#save axes
bc_axis.1 = ord$vectors[,1]
bc_axis.2 = ord$vectors[,2]
```

### Merge BC axis with merged_meta_r 
```{r}
bc_axis1.df <- data.frame(
  Subject.ID = names(bc_axis.1),
  bray_pcoa1 = as.numeric(bc_axis.1)
)

merged_meta_r <- merged_meta_r %>%
  dplyr::select(-any_of("bray_pcoa1")) %>%
  left_join(bc_axis1.df, by = "Subject.ID")

sum(is.na(merged_meta_r$bray_pcoa1))
head(merged_meta_r[, c("Subject.ID", "bray_pcoa1")])
all(bc_axis1.df$Subject.ID %in% merged_meta_r$Subject.ID)
```

### CLR, Aitchison PCA, PAM clusters

```{r}
library(zCompositions)
library(compositions)
library(cluster)


# Select k for PAM clustering
set.seed(123)
k.max <- 10

SIL <- sapply(2:k.max, function(i) {
  pam_fit <- pam(aitch, i, nstart = 50, diss = TRUE)
  mean(cluster::silhouette(pam_fit$clustering, dist = aitch)[, "sil_width"])
})

aitch_plot <- plot(
  2:k.max,
  SIL,
  type = "b",
  main = "Selecting k PAM clusters",
  ylab = "Mean Silhouette Width",
  xlab = "Clusters"
)


sil_df <- data.frame(
  Clusters = 2:k.max,
  Mean_Silhouette = SIL
)

p_sil <- ggplot(
  sil_df,
  aes(x = Clusters, y = Mean_Silhouette)
) +
  geom_line() +
  geom_point(size = 3) +
  theme_classic(base_size = 16) +
  theme(
    axis.text = element_text(size = 18, color = "black"),
    axis.title = element_text(size = 18, face = "bold"),
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA)
  ) +
  labs(
    x = "Clusters",
    y = "Mean Silhouette Width"
  )

p_sil

```

## Save SIL
```{r}
ggsave(
  filename = "Aitch_SIL_plot.pdf",
  plot = p_sil,
  path = fig_dir,
  width = 6,
  height = 5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)
```

# PAM clustering using k = 3 --> identified 3 clusters after updating samples
```{r}
set.seed(123)
pam_aitch <- pam(aitch, 3, nstart = 50, diss = TRUE)

cluster_df <- data.frame(
  Subject.ID = rownames(clr_mat),
  Cluster_Aitchison = factor(pam_aitch$clustering)
)

scores <- scores %>%
  dplyr::left_join(cluster_df, by = "Subject.ID")

# Axis labels
var_exp <- (pca1$sdev^2) / sum(pca1$sdev^2)

xlab <- paste0("PC1 [", round(100 * var_exp[1], 1), "%]")
ylab <- paste0("PC2 [", round(100 * var_exp[2], 1), "%]")

# Add Aitchison axes and clusters directly to merged_meta_r
merged_meta_r <- merged_meta_r %>%
  dplyr::select(
    -any_of(c(
      "PC1",
      "PC2",
      "Cluster_Aitchison"
    ))
  ) %>%
  dplyr::left_join(
    scores %>%
      dplyr::select(
        Subject.ID,
        PC1,
        PC2,
        Cluster_Aitchison
      ),
    by = "Subject.ID"
  )

# Keep CLR taxa table available for later taxa/heatmap analyses
clr_df <- as.data.frame(clr_mat)
clr_df$Subject.ID <- rownames(clr_df)

pain_cols_n <- c(
  "No pain (n = 6)" = "grey",
  "Mild pain (n = 8)" = "#FDBB84",
  "Moderate pain (n = 7)" = "#EF6548",
  "Severe pain (n = 9)" = "#990000"
)

merged_meta_r <- merged_meta_r %>%
  mutate(
    pain_severity_plot = factor(
      as.character(pain_severity),
      levels = c(
        "No pain (n = 6)",
        "Mild pain (n = 8)",
        "Moderate pain (n = 7)",
        "Severe pain (n = 9)"
      ),
      ordered = TRUE
    ),
    pain_severity_clean = factor(
      dplyr::case_when(
        as.character(pain_severity) == "No pain (n = 6)" ~ "No pain",
        as.character(pain_severity) == "Mild pain (n = 8)" ~ "Mild pain",
        as.character(pain_severity) == "Moderate pain (n = 7)" ~ "Moderate pain",
        as.character(pain_severity) == "Severe pain (n = 9)" ~ "Severe pain"
      ),
      levels = c("No pain", "Mild pain", "Moderate pain", "Severe pain"),
      ordered = TRUE
    )
  )

# merged_meta_r$Cluster_Aitchison <- factor(
#   merged_meta_r$Cluster_Aitchison,
#   levels = c(1, 2, 3),
#   labels = c("Cluster 1", "Cluster 2", "Cluster 3")
# )

# Plot Aitchison PCA by pain severity and cluster
aitch_plot <- ggplot(
  merged_meta_r,
  aes(
    x = PC1,
    y = PC2,
    color = pain_severity,
    shape = Cluster_Aitchison
  )
) +
  
  stat_ellipse(
    data = merged_meta_r,
    aes(group = Cluster_Aitchison),
    linewidth = 0.5,
    linetype = 2,
    #color = "black",
    type = "norm"
  ) +
  
  geom_point(size = 4) +
  scale_color_manual(values = pain_cols_n) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    plot.title = element_text(size = 24, hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(size = 22, hjust = 0.5),
    legend.title = element_text(size = 18, face = "bold"),
    legend.text = element_text(size = 18),
    legend.margin = margin(0, 0, 0, 0),
    
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank()
  ) +
  labs(
    title = "Aitchison Distance",
    x = xlab,
    y = ylab,
    color = "Pain Severity",
    shape = "Aitchison Cluster"
  )

aitch_plot
sum(is.na(merged_meta_r$PC1))
sum(is.na(merged_meta_r$Cluster_Aitchison))
all(scores$Subject.ID %in% merged_meta_r$Subject.ID)

```

### Save Clusters
```{r}
ggsave(
  filename = "Aitchison_pain.png",
  plot = aitch_plot,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  bg = "transparent"
)

ggsave(
  filename = "Aitchison_pain.pdf",
  plot = aitch_plot,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)

```

### DAA using ZicoSeq with Respect to pain severity 

```{r}
############################################################
# ZicoSeq DAA by categorical pain severity
############################################################
set.seed(123)

library(GUniFrac)
library(phyloseq)
library(tidyverse)
library(ggrepel)

# ----------------------------------------------------------
# 1. Prepare clean pain severity metadata
# ----------------------------------------------------------

merged_meta_r <- merged_meta_r %>%
  mutate(
    pain_severity_cat = case_when(
      grepl("^No pain", pain_severity) ~ "No pain",
      grepl("^Mild pain", pain_severity) ~ "Mild pain",
      grepl("^Moderate pain", pain_severity) ~ "Moderate pain",
      grepl("^Severe pain", pain_severity) ~ "Severe pain",
      TRUE ~ NA_character_
    ),
    pain_severity_cat = factor(
      pain_severity_cat,
      levels = c("No pain", "Mild pain", "Moderate pain", "Severe pain")
    )
  )

table(merged_meta_r$pain_severity_cat, useNA = "ifany")


# ----------------------------------------------------------
# 2. Build genus-level count matrix: taxa x samples
# ----------------------------------------------------------

otu_zico <- as(otu_table(merged_r_genus), "matrix")

if (!taxa_are_rows(merged_r_genus)) {
  otu_zico <- t(otu_zico)
}

storage.mode(otu_zico) <- "numeric"

# Preserve taxa names safely
taxa_ids <- taxa_names(merged_r_genus)

if (length(taxa_ids) == nrow(otu_zico)) {
  rownames(otu_zico) <- taxa_ids
} else {
  rownames(otu_zico) <- paste0("Taxon_", seq_len(nrow(otu_zico)))
}

rownames(otu_zico) <- make.unique(as.character(rownames(otu_zico)))


# ----------------------------------------------------------
# 3. Match metadata to OTU columns
# ----------------------------------------------------------

meta_zico <- merged_meta_r %>%
  filter(
    Subject.ID %in% colnames(otu_zico),
    !is.na(pain_severity_cat)
  ) %>%
  arrange(match(Subject.ID, colnames(otu_zico))) %>%
  as.data.frame()

rownames(meta_zico) <- meta_zico$Subject.ID

otu_zico <- otu_zico[, rownames(meta_zico), drop = FALSE]

# One prevalence filter only
otu_zico <- otu_zico[rowSums(otu_zico > 0) >= 3, , drop = FALSE]

# Sanity checks
stopifnot(nrow(meta_zico) > 0)
stopifnot(nrow(otu_zico) > 0)
stopifnot(ncol(otu_zico) == nrow(meta_zico))
stopifnot(all(colnames(otu_zico) == rownames(meta_zico)))


# ----------------------------------------------------------
# 4. Run ZicoSeq
# ----------------------------------------------------------

set.seed(123)

zico_pain_categorical <- ZicoSeq(
  meta.dat = meta_zico,
  feature.dat = otu_zico,
  grp.name = "pain_severity_cat",
  adj.name = NULL,
  feature.dat.type = "count",
  is.winsor = TRUE,
  outlier.pct = 0.03,
  winsor.end = "top",
  is.post.sample = FALSE,
  link.func = list(function(x) x^0.5),
  stats.combine.func = max,
  perm.no = 999,
  strata = NULL,
  ref.pct = 0.5,
  stage.no = 6,
  excl.pct = 0.2,
  is.fwer = FALSE,
  verbose = TRUE,
  return.feature.dat = TRUE
)

# ----------------------------------------------------------
# 5. Build clean ZicoSeq results table
# ----------------------------------------------------------

zico_results_categorical <- data.frame(
  Taxon = names(zico_pain_categorical$p.raw),
  p.raw = as.numeric(zico_pain_categorical$p.raw),
  p.adj.fdr = as.numeric(zico_pain_categorical$p.adj.fdr),
  R2 = as.numeric(zico_pain_categorical$R2)
) %>%
  arrange(p.adj.fdr)

# Add taxonomy
tax_df <- as.data.frame(tax_table(merged_r_genus))
tax_df$Taxon <- rownames(tax_df)

zico_results_categorical <- zico_results_categorical %>%
  left_join(tax_df, by = "Taxon")

# Pull significant taxa
sig_zico_taxa <- zico_results_categorical %>%
  filter(p.adj.fdr < 0.05)

sig_zico_taxa

# ----------------------------------------------------------
# 6. Generic ZicoSeq DAA volcano plot
# ----------------------------------------------------------

zico_plot_df <- zico_results_categorical %>%
  mutate(
    neg_log10_fdr = -log10(p.adj.fdr),
    Significant = ifelse(p.adj.fdr < 0.05, "FDR < 0.05", "Non-significant"),
    Genus_clean = case_when(
      Genus == "W5053" ~ "Clostridiales Family XI W5053",
      is.na(Genus) | Genus == "" ~ Taxon,
      TRUE ~ stringr::str_replace_all(Genus, "_", " ")
    ),
    Label = ifelse(
      p.adj.fdr < 0.05,
      paste0("italic(\"", Genus_clean, "\")"),
      NA_character_
    )
  )

zico_plot <- ggplot(
  zico_plot_df,
  aes(
    x = R2,
    y = neg_log10_fdr,
    color = Significant
  )
) +
  geom_point(size = 3, alpha = 0.8) +
  geom_hline(
    yintercept = -log10(0.05),
    linetype = "dashed",
    color = "black"
  ) +
  geom_text_repel(
    aes(label = Label),
    size = 8,
    parse = TRUE,
    max.overlaps = Inf,
    box.padding = 0.5,
    point.padding = 0.4,
    na.rm = TRUE,
    show.legend = FALSE
  ) +
  scale_color_manual(
    values = c(
      "Non-significant" = "grey75",
      "FDR < 0.05" = "#cb181d"
    )
  ) +
  theme_classic(base_size = 16) +
  theme(
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    #plot.title = element_text(size = 22, hjust = 0.5),
    plot.subtitle = element_text(size = 24, hjust = 0.5),
    legend.title = element_text(size = 18),
    legend.text = element_text(size = 18),
    legend.margin = margin(0, 0, 0, 0),
    
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank()
  ) +
  labs(
    x = expression(R^2),
    y = expression(-log[10]("FDR")),
    color = ""
  )
zico_plot
```

## Save DAA
```{r}
ggsave(
  filename = "zico_pain.png",
  plot = zico_plot,
  path = fig_dir,
  width = 9,
  height = 5.5,
  units = "in",
  bg = "transparent"
)

ggsave(
  filename = "zico_pain.pdf",
  plot = zico_plot,
  path = fig_dir,
  width = 9,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)
```

### Heatmap for taxa by pain severity]
i am unsure if we are using the right values, is CLR still appropriate even though zicoseq should use counts? 

R: CLR is good here. ZicoSeq as a function accepts counts, but it internally uses reference taxa to deal with compositionality, and a square root transform to normalize the count distribution (I added that).

Need to update boxplots to fix it not being mean of median...look at distribution of the medians for summary of CLR. This is showing distribution of medians, white=mean of the median...too confusing, update to be more straightforward.

```{r}
############################################################
# CLR table with original taxon IDs for ZicoSeq taxa
############################################################

clr_df_zico <- as.data.frame(clr_mat)

colnames(clr_df_zico) <- taxa_names(merged_r_genus)

clr_df_zico$Subject.ID <- rownames(clr_df_zico)

clr_meta <- clr_df_zico %>%
  left_join(
    merged_meta_r %>%
      dplyr::select(Subject.ID, pain_severity_cat),
    by = "Subject.ID"
  )

# Calculate median CLR per pain group
heat_df <- clr_meta %>%
  dplyr::select(
    Subject.ID,
    pain_severity_cat,
    all_of(sig_zico_taxa$Taxon)
  ) %>%
  pivot_longer(
    cols = all_of(sig_zico_taxa$Taxon),
    names_to = "Taxon",
    values_to = "CLR"
  ) %>%
  group_by(Taxon, pain_severity_cat) %>%
  summarise(
    Median_CLR = median(CLR, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(
    tax_df %>%
      dplyr::select(Taxon, Genus),
    by = "Taxon"
  ) %>%
  mutate(
    Genus_clean = ifelse(
      is.na(Genus) | Genus == "",
      Taxon,
      stringr::str_replace_all(Genus, "_", " ")
    )
  )

# Convert to matrix
heat_matrix <- heat_df %>%
  dplyr::select(
    Genus_clean,
    pain_severity_cat,
    Median_CLR
  ) %>%
  pivot_wider(
    names_from = pain_severity_cat,
    values_from = Median_CLR
  )

heat_mat <- as.data.frame(heat_matrix)
rownames(heat_mat) <- heat_mat$Genus_clean
heat_mat$Genus_clean <- NULL

heat_mat <- as.matrix(heat_mat)

# Plot heatmap
breaks <- seq(-1.5, 1.5, length.out = 101)

pheatmap::pheatmap(
  heat_mat,
  scale = "row",
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  color = colorRampPalette(c("#2166AC", "white", "#B2182B"))(100),
  breaks = breaks,
  fontsize_row = 10,
  main = "Pain severity-associated taxa",
  legend_breaks = c(-1.5, -1, 0, 1, 1.5),
  legend_labels = c(
    "Lower\n(-1.5)",
    "-1",
    "Mean\n(0)",
    "+1",
    "Higher\n(+1.5)"
  )
)

```

### Boxplots for significant taxa

```{r}
############################################################
# Individual boxplots for significant taxa
############################################################

set.seed(123)

library(dplyr)
library(tidyr)
library(ggplot2)
library(ggpubr)
library(purrr)

plot_df <- clr_meta %>%
  dplyr::select(
    Subject.ID,
    pain_severity_cat,
    all_of(sig_zico_taxa$Taxon)
  ) %>%
  pivot_longer(
    cols = all_of(sig_zico_taxa$Taxon),
    names_to = "Taxon",
    values_to = "CLR"
  ) %>%
  left_join(
    tax_df %>%
      dplyr::select(Taxon, Genus),
    by = "Taxon"
  ) %>%
  mutate(
    Genus_clean = case_when(
      Genus == "W5053" ~ "Clostridiales Family XI W5053",
      is.na(Genus) | Genus == "" ~ Taxon,
      TRUE ~ stringr::str_replace_all(Genus, "_", " ")
    ),
    
    pain_severity_cat = factor(
      pain_severity_cat,
      levels = c(
        "No pain",
        "Mild pain",
        "Moderate pain",
        "Severe pain"
      ),
      labels = c(
        "No pain\n(n = 6)",
        "Mild pain\n(n = 8)",
        "Moderate pain\n(n = 7)",
        "Severe pain\n(n = 9)"
      )
    )
  ) %>%
  filter(!is.na(pain_severity_cat))

comparisons <- list(
  c("No pain\n(n = 6)", "Mild pain\n(n = 8)"),
  c("No pain\n(n = 6)", "Moderate pain\n(n = 7)"),
  c("No pain\n(n = 6)", "Severe pain\n(n = 9)"),
  c("Mild pain\n(n = 8)", "Moderate pain\n(n = 7)"),
  c("Mild pain\n(n = 8)", "Severe pain\n(n = 9)"),
  c("Moderate pain\n(n = 7)", "Severe pain\n(n = 9)")
)

############################################################
# Function for significant comparisons
############################################################

get_sig_comparisons <- function(df) {
  
  res <- purrr::map_dfr(comparisons, function(comp) {
    
    temp <- df %>%
      filter(pain_severity_cat %in% comp) %>%
      droplevels()
    
    if(length(unique(temp$pain_severity_cat)) < 2) {
      return(NULL)
    }
    
    test <- wilcox.test(CLR ~ pain_severity_cat, data = temp)
    
    tibble(
      group1 = comp[1],
      group2 = comp[2],
      p = test$p.value
    )
  }) %>%
    mutate(
      p.adj = p.adjust(p, method = "fdr"),
      p.signif = case_when(
        p.adj <= 0.001 ~ "***",
        p.adj <= 0.01 ~ "**",
        p.adj <= 0.05 ~ "*",
        TRUE ~ "ns"
      )
    ) %>%
    filter(p.adj <= 0.05)
  
  return(res)
}

############################################################
# Function to generate one plot
############################################################

make_taxon_plot <- function(genus_name) {
  
  df_taxon <- plot_df %>%
    filter(Genus_clean == genus_name)
  
  sig_tests <- get_sig_comparisons(df_taxon)
  
  y_max <- max(df_taxon$CLR, na.rm = TRUE)
  y_min <- min(df_taxon$CLR, na.rm = TRUE)
  step <- (y_max - y_min) * 0.18
  scale_y_continuous(
    expand = expansion(mult = c(0.05, 0.35))
  )
  
  if(nrow(sig_tests) > 0) {
    sig_tests <- sig_tests %>%
      mutate(
        y.position = y_max + step * row_number()
      )
  }
  
  plot_title <- if(genus_name == "W5053") {
    bquote(italic("Clostridiales Family XI W5053"))
  } else {
    bquote(italic(.(genus_name)))
  }
  
  p <- ggplot(
    df_taxon,
    aes(
      x = pain_severity_cat,
      y = CLR,
      fill = pain_severity_cat
    )
  ) +
    geom_boxplot(
      outlier.shape = NA,
      alpha = 0.8,
      width = 0.65
    ) +
    geom_jitter(
      width = 0.12,
      size = 3,
      color = "black"
    ) +
    stat_pvalue_manual(
      sig_tests,
      label = "p.signif",
      tip.length = 0.01,
      bracket.size = 0.7,
      size = 12
    ) +
    scale_fill_manual(
      values = c(
        "No pain\n(n = 6)" = "grey",
        "Mild pain\n(n = 8)" = "#FDBB84",
        "Moderate pain\n(n = 7)" = "#EF6548",
        "Severe pain\n(n = 9)" = "#990000"
      ),
      drop = FALSE
    ) +
    theme_classic(base_size = 18) +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 45, hjust = 1, size = 20, color = "black", face = "bold"),
          axis.text.y = element_text(size = 20, color = "black"),
          axis.title.x = element_text(size = 18),
          axis.title.y = element_text(size = 20,  face = "bold"),
          plot.title = element_text(size = 26, hjust = 0.5, face = "bold"),
          #plot.subtitle = element_text(size = 18, hjust = 0.5),
          legend.title = element_text(size = 18),
          legend.text = element_text(size = 18),
          legend.margin = margin(0, 0, 0, 0),
          
          panel.background = element_rect(fill = "transparent", colour = NA),
          plot.background = element_rect(fill = "transparent", colour = NA),
          
          legend.background = element_blank(),
          legend.box.background = element_blank(),
          legend.key = element_blank()
    ) +
    labs(
      title = plot_title,
      x = NULL,
      y = "CLR Abundance"
    )
  
  return(p)
}

############################################################
# Generate plots
############################################################

taxon_plots <- purrr::map(
  unique(plot_df$Genus_clean),
  make_taxon_plot
)


names(taxon_plots) <- unique(plot_df$Genus_clean)
taxon_plots$Ezakiella
taxon_plots$`Clostridiales Family XI W5053`
```

```{r}

############################################################
# Save plots
############################################################
fig_dir <- "C:/Users/m298096/OneDrive - Mayo Clinic/1-Thesis_Research/1-Manuscripts/Microbiome Manuscript/MKB figs/[CURRENT] Pain plots/Rectal  Plots"

ggsave(
  filename = "W5053_pain_boxplot.png",
  plot = taxon_plots[["Clostridiales Family XI W5053"]],
  path = fig_dir,
  width = 8,
  height = 7,
  units = "in",
  dpi = 600,
  bg = "transparent"
)

ggsave(
  filename = "W5053_pain_boxplot.pdf",
  plot = taxon_plots[["Clostridiales Family XI W5053"]],
  path = fig_dir,
  width = 8,
  height = 7,
  units = "in",
  device = "pdf",
  bg = "transparent"
)

ggsave(
  filename = "Ezakiella_pain_boxplot.png",
  plot = taxon_plots$Ezakiella,
  path = fig_dir,
  width = 8,
  height = 7,
  units = "in",
  dpi = 600,
  bg = "transparent"
)

ggsave(
  filename = "Ezakiella_pain_boxplot.pdf",
  plot = taxon_plots$Ezakiella,
  path = fig_dir,
  width = 8,
  height = 7,
  units = "in",
  device = "pdf",
  bg = "transparent"
)
```


### Fisher's test for cluster based on pain severity
```{r}
fisher.test(
  table(
    merged_meta_r$Cluster_Aitchison,
    merged_meta_r$pain_severity
  )
)

tab <- table(
  merged_meta_r$Cluster_Aitchison,
  merged_meta_r$pain_severity
)

fisher.test(tab)

pairwise_results <- list()

pain_levels <- levels(merged_meta_r$pain_severity)

for(i in 1:(length(pain_levels)-1)) {
  for(j in (i+1):length(pain_levels)) {
    
    temp <- merged_meta_r %>%
      filter(
        pain_severity %in% c(
          pain_levels[i],
          pain_levels[j]
        )
      )
    
    tab <- table(
      temp$Cluster_Aitchison,
      temp$pain_severity
    )
    
    res <- fisher.test(tab)
    
    pairwise_results[[paste(
      pain_levels[i],
      "vs",
      pain_levels[j]
    )]] <- res$p.value
  }
}

pairwise_results

p.adjust(
  unlist(pairwise_results),
  method = "fdr"
)

```

## Graph cluster counts and Fisher's results
```{r}
cluster_counts <- merged_meta_r %>%
  mutate(
    pain_severity_plot = case_when(
      grepl("^No pain", as.character(pain_severity)) ~ "No pain (n = 6)",
      grepl("^Mild pain", as.character(pain_severity)) ~ "Mild pain (n = 8)",
      grepl("^Moderate pain", as.character(pain_severity)) ~ "Moderate pain (n = 7)",
      grepl("^Severe pain", as.character(pain_severity)) ~ "Severe pain (n = 9)",
      TRUE ~ NA_character_
    ),
    pain_severity_plot = factor(
      pain_severity_plot,
      levels = c(
        "No pain (n = 6)",
        "Mild pain (n = 8)",
        "Moderate pain (n = 7)",
        "Severe pain (n = 9)"
      )
    ),
    Cluster_Aitchison = factor(
      Cluster_Aitchison,
      levels = c(1, 2, 3),
      labels = c("Cluster 1", "Cluster 2", "Cluster 3")
    )
  ) %>%
  filter(!is.na(pain_severity_plot)) %>%
  count(Cluster_Aitchison, pain_severity_plot, .drop = FALSE) %>%
  group_by(Cluster_Aitchison) %>%
  mutate(
    prop = n / sum(n),
    label = ifelse(n == 0, "", as.character(n))
  ) %>%
  ungroup()

p.adjust(
  unlist(pairwise_results),
  method = "fdr"
)


cluster_counts <- merged_meta_r %>%
  mutate(
    pain_severity_plot = case_when(
      grepl("^No pain", as.character(pain_severity)) ~ "No pain",
      grepl("^Mild pain", as.character(pain_severity)) ~ "Mild",
      grepl("^Moderate pain", as.character(pain_severity)) ~ "Moderate",
      grepl("^Severe pain", as.character(pain_severity)) ~ "Severe",
      TRUE ~ NA_character_
    ),
    pain_severity_plot = factor(
      pain_severity_plot,
      levels = c("No pain", "Mild", "Moderate", "Severe")
    ),
    Cluster_Aitchison = factor(
      Cluster_Aitchison,
      levels = c(1, 2, 3),
      labels = c("Cluster 1", "Cluster 2", "Cluster 3")
    )
  ) %>%
  filter(!is.na(pain_severity_plot), !is.na(Cluster_Aitchison)) %>%
  count(Cluster_Aitchison, pain_severity_plot, .drop = FALSE) %>%
  mutate(label = ifelse(n == 0, "", as.character(n)))

clusters_pain_counts <- ggplot(
  cluster_counts,
  aes(
    x = Cluster_Aitchison,
    y = n,
    fill = pain_severity_plot
  )
) +
  geom_col(width = 0.85, color = "white") +
  geom_text(
    aes(label = label),
    position = position_stack(vjust = 0.5),
    size = 6,
    color = "black"
  ) +
  scale_fill_manual(
    values = c(
      "No pain" = "#737373",
      "Mild" = "#FDBB84",
      "Moderate" = "#EF6548",
      "Severe" = "#990000"
    ),
    drop = FALSE
  ) +
  scale_y_continuous(
    breaks = seq(0, 14, by = 2),
    expand = expansion(mult = c(0, 0.05))
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 20, face = "bold"),
    legend.title = element_text(size = 18, face = "bold"),
    legend.text = element_text(size = 16),
    legend.key = element_blank(),
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA)
  ) +
  labs(
    x = NULL,
    y = "Number of Participants",
    fill = "Pain Severity"
  )

clusters_pain_counts

## Add fisher's p-value to the plot
fisher_p <- fisher.test(
  table(
    merged_meta_r$Cluster_Aitchison,
    merged_meta_r$pain_severity
  )
)$p.value

clusters_pain_counts <- clusters_pain_counts +
  labs(
    subtitle = bquote("Fisher's Exact " * italic(p) == .(signif(fisher_p, 3)))
  ) +
  theme(
    plot.subtitle = element_text(size = 20, hjust = 0.5, face = "bold")
  )

clusters_pain_counts

## Pairwise comparisons
pairwise_df <- tibble::tibble(
  comparison = names(pairwise_results),
  p_value = unlist(pairwise_results),
  fdr = p.adjust(unlist(pairwise_results), method = "fdr")
) %>%
  dplyr::mutate(
    p_label = paste0("p = ", signif(p_value, 3)),
    fdr_label = paste0("FDR = ", signif(fdr, 3))
  )

pairwise_df


```
# Save 
```{r}
ggsave(
  filename = "aitch_pain_clusters.png",
  plot = clusters_pain_counts,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  dpi = 600,
  bg = "transparent"
)

ggsave(
  filename = "aitch_pain_clusters.pdf",
  plot = clusters_pain_counts,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)
```

### Secondary cluster analayses...zicoseq 

```{r, eval =F}
############################################################
# Secondary analysis: characterize Aitchison clusters
# Taxa + predicted functional ecology
############################################################

library(GUniFrac)
library(microeco)
library(phyloseq)
library(tidyverse)
library(ggrepel)
library(pheatmap)
library(ggpubr)

# Make sure cluster is a factor
merged_meta_r <- merged_meta_r %>%
  mutate(
    Cluster_Aitchison = factor(Cluster_Aitchison)
  )

############################################################
# Part 1. ZicoSeq DAA by Aitchison cluster
############################################################

# Genus-level count matrix
otu_cluster <- as.data.frame(otu_table(merged_r_genus))

# ZicoSeq expects taxa x samples
if (!taxa_are_rows(merged_r_genus)) {
  otu_cluster <- t(otu_cluster)
}

otu_cluster <- as.matrix(otu_cluster)
storage.mode(otu_cluster) <- "numeric"

# Match metadata to OTU columns
meta_cluster <- merged_meta_r %>%
  filter(
    Subject.ID %in% colnames(otu_cluster),
    !is.na(Cluster_Aitchison)
  ) %>%
  arrange(match(Subject.ID, colnames(otu_cluster))) %>%
  as.data.frame()

rownames(meta_cluster) <- meta_cluster$Subject.ID

otu_cluster <- otu_cluster[, meta_cluster$Subject.ID]

# Optional analysis-specific filter
otu_cluster <- otu_cluster[rowSums(otu_cluster > 0) >= 3, ]

# Checks
dim(otu_cluster)
nrow(meta_cluster)
all(colnames(otu_cluster) == rownames(meta_cluster))

set.seed(123)

zico_cluster <- ZicoSeq(
  meta.dat = meta_cluster,
  feature.dat = otu_cluster,
  grp.name = "Cluster_Aitchison",
  adj.name = NULL,
  feature.dat.type = "count",
  is.winsor = TRUE,
  outlier.pct = 0.03,
  winsor.end = "top",
  is.post.sample = FALSE,
  link.func = list(function(x) x^0.5),
  stats.combine.func = max,
  perm.no = 999,
  strata = NULL,
  ref.pct = 0.5,
  stage.no = 6,
  excl.pct = 0.2,
  is.fwer = FALSE,
  verbose = TRUE,
  return.feature.dat = TRUE
)

# Results table
zico_cluster_results <- data.frame(
  Taxon = names(zico_cluster$p.raw),
  p.raw = as.numeric(zico_cluster$p.raw),
  p.adj.fdr = as.numeric(zico_cluster$p.adj.fdr),
  R2 = as.numeric(zico_cluster$R2)
) %>%
  arrange(p.adj.fdr)

# Add taxonomy
tax_df_cluster <- as.data.frame(tax_table(merged_r_genus))
tax_df_cluster$Taxon <- rownames(tax_df_cluster)

zico_cluster_results <- zico_cluster_results %>%
  left_join(tax_df_cluster, by = "Taxon") %>%
  mutate(
    Genus_clean = ifelse(
      is.na(Genus) | Genus == "",
      Taxon,
      stringr::str_replace_all(Genus, "_", " ")
    ),
    Genus_clean = case_when(
      Taxon == "213ebca51a1d1b76ccd1456820b11a96548261a1" ~ "Eggerthellaceae unclassified",
      Taxon == "d7d0c7358e32837b6121cfd76b27f13f2a1edb75" ~ "Enterobacteriaceae unclassified",
      TRUE ~ Genus_clean
    )
  )
sig_zico_cluster <- zico_cluster_results %>%
  filter(p.adj.fdr < 0.05) 

sig_zico_cluster


############################################################
# Part 2. Cluster-defining taxa heatmap
############################################################

# CLR matrix with original taxon IDs
clr_cluster <- as.data.frame(clr_mat)
colnames(clr_cluster) <- taxa_names(merged_r_genus)
clr_cluster$Subject.ID <- rownames(clr_cluster)

clr_cluster_meta <- clr_cluster %>%
  left_join(
    merged_meta_r %>%
      dplyr::select(Subject.ID, Cluster_Aitchison),
    by = "Subject.ID"
  )

############################################################
# Part 3. Identify which Cluster each sig taxa is highest in
############################################################
if (nrow(sig_zico_cluster) > 0) {
  
  cluster_taxa_medians <- clr_cluster_meta %>%
    dplyr::select(
      Subject.ID,
      Cluster_Aitchison,
      all_of(sig_zico_cluster$Taxon)
    ) %>%
    pivot_longer(
      cols = all_of(sig_zico_cluster$Taxon),
      names_to = "Taxon",
      values_to = "CLR"
    ) %>%
    group_by(Taxon, Cluster_Aitchison) %>%
    summarise(
      Median_CLR = median(CLR, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    left_join(
      tax_df_cluster %>%
        dplyr::select(Taxon, Genus),
      by = "Taxon"
    ) %>%
    mutate(
      Genus_clean = ifelse(
        is.na(Genus) | Genus == "",
        Taxon,
        stringr::str_replace_all(Genus, "_", " ")
      )
    )
}

cluster_taxa_direction <- cluster_taxa_medians %>%
  group_by(Taxon) %>%
  mutate(
    Highest_cluster = Cluster_Aitchison[which.max(Median_CLR)],
    Highest_median_CLR = max(Median_CLR, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  arrange(Highest_cluster, Genus_clean, Cluster_Aitchison)

cluster_taxa_direction

cluster_taxa_summary <- cluster_taxa_medians %>%
  group_by(Taxon, Genus_clean) %>%
  slice_max(
    order_by = Median_CLR,
    n = 1,
    with_ties = FALSE
  ) %>%
  ungroup() %>%
  rename(
    Highest_cluster = Cluster_Aitchison,
    Highest_median_CLR = Median_CLR
  ) %>%
  arrange(Highest_cluster, Genus_clean)

cluster_taxa_summary

############################################################
# Part 4. Generic DAA volcano plot for clusters
############################################################
cluster_taxa_medians <- cluster_taxa_medians %>%
  dplyr::mutate(
    Genus_clean = dplyr::case_when(
      Taxon == "213ebca51a1d1b76ccd1456820b11a96548261a1" ~ "Eggerthellaceae unclassified",
      Taxon == "d7d0c7358e32837b6121cfd76b27f13f2a1edb75" ~ "Enterobacteriaceae unclassified",
      TRUE ~ Genus_clean
    )
  )

cluster_taxa_medians <- cluster_taxa_medians %>%
  mutate(
    Genus_label = case_when(
      Genus_clean == "Eggerthellaceae unclassified" ~
        "italic('Eggerthellaceae')~'unclassified'",

      Genus_clean == "Enterobacteriaceae unclassified" ~
        "italic('Enterobacteriaceae')~'unclassified'",

      TRUE ~ paste0("italic('", Genus_clean, "')")
    )
  )

cluster_taxa_plot_df <- cluster_taxa_medians %>%
  group_by(Taxon, Genus_clean) %>%
  mutate(
    Highest_cluster = Cluster_Aitchison[which.max(Median_CLR)]
  ) %>%
  ungroup() %>%
  mutate(
    Highest_cluster = factor(Highest_cluster),
    Cluster_Aitchison = factor(Cluster_Aitchison),
    Genus_clean = forcats::fct_reorder(Genus_clean, as.numeric(Highest_cluster))
  )

cluster_taxa_plot <- ggplot(
  cluster_taxa_plot_df,
  aes(
    x = Cluster_Aitchison,
    y = Genus_clean,
    fill = Median_CLR
  )
) +
  geom_tile(color = "white", linewidth = 0.6) +
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0,
    name = "Median CLR"
  ) +
  scale_y_discrete(
  labels = function(x) parse(text =
    cluster_taxa_plot_df$Genus_label[
      match(x, cluster_taxa_plot_df$Genus_clean)
    ]
  )
) +
  theme_classic(base_size = 16) +
  labs(
    x = "Aitchison cluster",
    y = NULL,
    title = "Differentially Abundant Taxa by Cluster",
    fill = "Median CLR"
  ) +
  theme(
    axis.text.y = element_text(size = 12, color = "black"),
    axis.text.x = element_text(size = 14),
    axis.title.x = element_text(size = 16),
    plot.title = element_text(size = 18, hjust = 0.5, face = "bold"),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  )
cluster_taxa_plot <- cluster_taxa_plot +
  labs(
    x = NULL,
    title = NULL
  ) +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 16, color = "black"),
    legend.title = element_text(size = 18, face = "bold"),
    legend.text = element_text(size = 16),
    legend.key.height = unit(1.2, "cm"),
    legend.key.width = unit(0.35, "cm"),
    legend.margin = margin(0, 0, 0, 0),
    plot.margin = margin(5, 5, 5, 5),
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    legend.background = element_rect(fill = "transparent", colour = NA)
  )
cluster_taxa_plot

zico_clusters_df <- zico_cluster_results %>%
  mutate(
    neg_log10_fdr = -log10(p.adj.fdr),
    Significant = ifelse(p.adj.fdr < 0.05, "FDR < 0.05", "Non-significant"),
    Genus_clean = case_when(
      Genus == "W5053" ~ "Clostridiales Family XI W5053",
      Taxon == "213ebca51a1d1b76ccd1456820b11a96548261a1" ~ "Eggerthellaceae unclassified",
      Taxon == "d7d0c7358e32837b6121cfd76b27f13f2a1edb75" ~ "Enterobacteriaceae unclassified",
      is.na(Genus) | Genus == "" ~ Taxon,
      TRUE ~ stringr::str_replace_all(Genus, "_", " ")
    ),
    Label = ifelse(
      p.adj.fdr < 0.05,
      paste0("italic(\"", Genus_clean, "\")"),
      NA_character_
    )
  )


zico_plot_clusters <- ggplot(
  zico_clusters_df,
  aes(
    x = R2,
    y = neg_log10_fdr,
    color = Significant
  )
) +
  geom_point(size = 3, alpha = 0.8) +
  geom_hline(
    yintercept = -log10(0.05),
    linetype = "dashed",
    color = "black"
  ) +
  geom_text_repel(
    aes(label = Label),
    size = 7,
    parse = TRUE,
    max.overlaps = Inf,
    box.padding = 1.5,
    point.padding = 1.2,
    force = 20,
    force_pull = 0.02,
    nudge_y = 0.22,
    max.iter = 50000,
    max.time = 10,
    min.segment.length = 0,
    segment.color = "grey50",
    segment.size = 0.6,
    direction = "both",
    na.rm = TRUE,
    show.legend = FALSE
  ) +
  scale_color_manual(
    values = c(
      "Non-significant" = "grey75",
      "FDR < 0.05" = "#cb181d"
    )
  ) +
  scale_y_continuous(
  limits = c(0, 4),
  expand = expansion(mult = c(0.02, 0.2))
) +
  #scale_x_continuous(
  #limits = c(0, 0.6),
  #expand = expansion(mult = c(0.02, 0.05))
#) #+
  theme_classic(base_size = 16) +
  theme(
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    #plot.title = element_text(size = 22, hjust = 0.5),
    plot.subtitle = element_text(size = 24, hjust = 0.5),
    legend.title = element_text(size = 18),
    legend.text = element_text(size = 18),
    legend.margin = margin(0, 0, 0, 0),
    
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank()
  ) +
  labs(
    x = expression(R^2),
    y = expression(-log[10]("FDR")),
    color = ""
  )


zico_plot_clusters
```

```{r}
ggsave(
  filename = "cluster_taxa_plot.pdf",
  plot = cluster_taxa_plot,
  path = fig_dir,
  width = 9,
  height = 6.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)

ggsave(
  filename = "zico_plot_clusters.pdf",
  plot = zico_plot_clusters,
  path = fig_dir,
  width = 12,
  height = 8,
  units = "in",
  device = "pdf",
  bg = "transparent"
)

```

```{r}
############################################################
# 5. Pairwise comparisons between clusters for significant taxa
############################################################

cluster_taxa_pairwise <- cluster_taxa_plot_df %>%
  dplyr::select(
    Taxon,
    Genus_clean
  ) %>%
  dplyr::distinct() %>%
  purrr::pmap_dfr(
    function(Taxon, Genus_clean) {
      
      temp <- clr_cluster_meta %>%
        dplyr::select(
          Subject.ID,
          Cluster_Aitchison,
          all_of(Taxon)
        ) %>%
        dplyr::rename(
          CLR = all_of(Taxon)
        ) %>%
        dplyr::filter(
          !is.na(Cluster_Aitchison),
          !is.na(CLR)
        )
      
      test <- pairwise.wilcox.test(
        x = temp$CLR,
        g = temp$Cluster_Aitchison,
        p.adjust.method = "fdr"
      )
      
      as.data.frame(as.table(test$p.value)) %>%
        dplyr::filter(!is.na(Freq)) %>%
        dplyr::mutate(
          Taxon = Taxon,
          Genus_clean = Genus_clean,
          Comparison = paste(Var1, "vs", Var2),
          Pairwise_FDR = Freq
        ) %>%
        dplyr::select(
          Taxon,
          Genus_clean,
          Comparison,
          Pairwise_FDR
        )
    }
  ) %>%
  dplyr::arrange(
    Genus_clean,
    Pairwise_FDR
  )

cluster_taxa_pairwise

cluster_taxa_pairwise_direction <- cluster_taxa_pairwise %>%
  tidyr::separate(
    Comparison,
    into = c("Cluster_A", "Cluster_B"),
    sep = " vs ",
    remove = FALSE
  ) %>%
  dplyr::left_join(
    cluster_taxa_medians %>%
      dplyr::rename(
        Cluster_A = Cluster_Aitchison,
        Median_CLR_A = Median_CLR
      ) %>%
      dplyr::select(Taxon, Cluster_A, Median_CLR_A),
    by = c("Taxon", "Cluster_A")
  ) %>%
  dplyr::left_join(
    cluster_taxa_medians %>%
      dplyr::rename(
        Cluster_B = Cluster_Aitchison,
        Median_CLR_B = Median_CLR
      ) %>%
      dplyr::select(Taxon, Cluster_B, Median_CLR_B),
    by = c("Taxon", "Cluster_B")
  ) %>%
  dplyr::mutate(
    Higher_cluster = dplyr::case_when(
      Median_CLR_A > Median_CLR_B ~ Cluster_A,
      Median_CLR_B > Median_CLR_A ~ Cluster_B,
      TRUE ~ "Tie"
    )
  ) %>%
  dplyr::arrange(
    Genus_clean,
    Pairwise_FDR
  )

cluster_taxa_pairwise_direction

cluster_taxa_pairwise_direction %>%
  dplyr::filter(Pairwise_FDR < 0.05)




############################################################
# 6. Plot cluster pairwise comparisons for significant taxa
############################################################

cluster_taxa_boxplot_df <- clr_cluster_meta %>%
  dplyr::select(
    Subject.ID,
    Cluster_Aitchison,
    all_of(unique(sig_zico_cluster$Taxon))
  ) %>%
  tidyr::pivot_longer(
    cols = all_of(unique(sig_zico_cluster$Taxon)),
    names_to = "Taxon",
    values_to = "CLR"
  ) %>%
  dplyr::left_join(
    sig_zico_cluster %>%
      dplyr::select(Taxon, Genus_clean),
    by = "Taxon"
  ) %>%
  dplyr::filter(
    !is.na(Cluster_Aitchison),
    !is.na(CLR)
  ) %>%
  dplyr::mutate(
    Cluster_Aitchison = factor(
      Cluster_Aitchison,
      levels = c("1", "2", "3")
    )
  )


plot_cluster_taxon <- function(taxon_name) {
  
  temp_plot <- cluster_taxa_boxplot_df %>%
    dplyr::filter(Genus_clean == taxon_name)
  
  temp_stats <- cluster_taxa_pairwise_direction %>%
    dplyr::filter(
      Genus_clean == taxon_name,
      Pairwise_FDR < 0.05
    ) %>%
    dplyr::mutate(
      group1 = Cluster_A,
      group2 = Cluster_B,
      label = dplyr::case_when(
        Pairwise_FDR < 0.001 ~ "***",
        Pairwise_FDR < 0.01 ~ "**",
        Pairwise_FDR < 0.05 ~ "*",
        TRUE ~ "ns"
      ),
      y.position = max(temp_plot$CLR, na.rm = TRUE) +
        dplyr::row_number() * 0.8
    )
  
  ggplot(
    temp_plot,
    aes(
      x = Cluster_Aitchison,
      y = CLR,
      fill = Cluster_Aitchison
    )
  ) +
    geom_boxplot(
      width = 0.65,
      outlier.shape = NA,
      color = "black",
      linewidth = 0.8
    ) +
    geom_jitter(
      width = 0.12,
      size = 3,
      color = "black",
      alpha = 0.9
    ) +
    ggpubr::stat_pvalue_manual(
      temp_stats,
      label = "label",
      y.position = "y.position",
      tip.length = 0.01,
      bracket.size = 0.8,
      size = 12,
      face = "bold"
    ) +
    scale_fill_manual(
      values = c(
        "1" = "#9E2F2F",
        "2" = "#FDBB84",
        "3" = "#CB181D"
      )
    ) +
    theme_classic(base_size = 18) +
    labs(
      x = "Aitchison Cluster",
      y = "CLR Abundance",
      title = taxon_name
    ) +
    theme(
      plot.title = element_text(
        face = "italic",
        size = 26,
        hjust = 0.5
      ),
      axis.text.x = element_text(
        size = 18
      ),
      axis.text.y = element_text(
        size = 18
      ),
      axis.title.x = element_text(size = 22, face = "bold"),
      axis.title.y = element_text(size = 22, face = "bold"),
      legend.position = "none",
      panel.background = element_rect(fill = "transparent", colour = NA),
      plot.background = element_rect(fill = "transparent", colour = NA)
  )
    
}

ezakiella_plot <- plot_cluster_taxon("Ezakiella")
ezakiella_plot
S5_plot <- plot_cluster_taxon("S5-A14a")
S5_plot
pepto_plot <- plot_cluster_taxon("Peptoniphilus")
pepto_plot
peptostrepto_plot <- plot_cluster_taxon("Peptostreptococcus")
peptostrepto_plot


plot_cluster_taxon("Peptostreptococcus")
plot_cluster_taxon("Ruminiclostridium 6")
plot_cluster_taxon("Intestinibacter")
plot_cluster_taxon("Sellimonas")
plot_cluster_taxon("Eggerthellaceae unclassified")
plot_cluster_taxon("Enterobacteriaceae unclassified")
plot_cluster_taxon("Adlercreutzia")
plot_cluster_taxon("Anaerosporobacter")
plot_cluster_taxon("Christensenellaceae R-7 group")
plot_cluster_taxon("Lachnoclostridium")
plot_cluster_taxon("Peptoniphilus")
plot_cluster_taxon("Ruminococcaceae UCG-002")
plot_cluster_taxon("Ruminococcus 1")
plot_cluster_taxon("Sutterella")

```


## Save significant taxa
```{r}
ggsave(
  filename = "ezakiella_plot.pdf",
  plot = ezakiella_plot,
  path = fig_dir,
  width = 8,
  height = 6,
  units = "in",
  device = "pdf",
  bg = "transparent"
)

ggsave(
  filename = "S5A14a_plot.pdf",
  plot = S5_plot,
  path = fig_dir,
  width = 8,
  height = 6,
  units = "in",
  device = "pdf",
  bg = "transparent"
)


ggsave(
  filename = "pepto_plot.pdf",
  plot = pepto_plot,
  path = fig_dir,
  width = 8,
  height = 6,
  units = "in",
  device = "pdf",
  bg = "transparent"
)

ggsave(
  filename = "peptostrepto_plot.pdf",
  plot = peptostrepto_plot,
  path = fig_dir,
  width = 8,
  height = 6,
  units = "in",
  device = "pdf",
  bg = "transparent"
)
```

### Network analyses

```{r}
library(SpiecEasi)
set.seed(123)
#get counts
sparcc_tab = merged_r |> tax_glom(taxrank = "Genus") |> tax_filter(min_prevalence = 0.2) #|> tax_select("unclassified", deselect = T)

sparcc_counts = otu_table(sparcc_tab)
sparcc_ra = sparcc_counts / rowSums(sparcc_counts) #rel abundance

#run sparcc
sparcc_res = sparcc(sparcc_ra)

#rename
rownames(sparcc_res$Cor) = colnames(sparcc_counts)
colnames(sparcc_res$Cor) = colnames(sparcc_counts)
rownames(sparcc_res$Cov) = colnames(sparcc_counts)
colnames(sparcc_res$Cov) = colnames(sparcc_counts)

sparcc_cor = sparcc_res$Cor

#re rename
sum(rownames(sparcc_cor) != rownames(tax_table(sparcc_tab))) #0 means names match
rownames(sparcc_cor) = tax_table(sparcc_tab)[,"Genus"]
colnames(sparcc_cor) = tax_table(sparcc_tab)[,"Genus"]

```


```{r}
get_lower_tri<-function(cormat){
  cormat[upper.tri(cormat)] <- NA
  return(cormat)
}
# Get upper triangle of the correlation matrix
get_upper_tri <- function(cormat){
  cormat[lower.tri(cormat)]<- NA
  return(cormat)
}

reorder_cormat <- function(cormat){
  # Use correlation between variables as distance
  dd <- as.dist((1-cormat)/2)
  hc <- hclust(dd)
  cormat <-cormat[hc$order, hc$order]
}

reorder_cor_and_p <- function(cormat, pmat){
  dd <- as.dist((1-cormat)/2)
  hc <- hclust(dd)
  cormat <-cormat[hc$order, hc$order]
  pmat <- pmat[hc$order, hc$order]
  list(r = cormat, p = pmat)
}

```


#### CLR and spearman correlations on pain subset

Some taxa pairs looked nonlinear, so I switched to spearman.

```{r}
library(psych) #remotes::install_version("psych", version = "2.5.3")

set.seed(123)
#get counts
sparcc_tab = merged_r |> tax_glom(taxrank = "Genus") |> tax_filter(min_prevalence = 0.2) #|> tax_select("unclassified", deselect = T)

sparcc_counts = otu_table(sparcc_tab)
sparcc_ra = sparcc_counts / rowSums(sparcc_counts) #rel abundance



#plot some rels to check assumptions
dfClr = as.data.frame(clr(sparcc_counts + 1))
subs = dfClr[, sample(colnames(dfClr), 10)]
pairs(subs)


MtxClr <- as.matrix(clr(sparcc_counts))


spearCorTestClr <- corr.test(MtxClr, method = "spearman", adjust = "none")
spearCorClr <- spearCorTestClr$r
spearPClr <- spearCorTestClr$p


sum(rownames(spearCorClr) != rownames(tax_table(sparcc_tab))) #0 means names match

# must comment out this renaming block to determine which asvs are unclassified and significant
rownames(spearCorClr) = tax_table(sparcc_tab)[,"Genus"]
colnames(spearCorClr) = tax_table(sparcc_tab)[,"Genus"]

rownames(spearPClr) = tax_table(sparcc_tab)[,"Genus"]
colnames(spearPClr) = tax_table(sparcc_tab)[,"Genus"]


reordered_all_Clr <- reorder_cor_and_p(spearCorClr, spearPClr)
reordered_spearCor_Clr <- reordered_all_Clr$r
reordered_spearP_Clr <- reordered_all_Clr$p


spearCor_processed_Clr <- reordered_spearCor_Clr  %>% get_upper_tri() %>% reshape2::melt() %>% na.omit() %>% dplyr::rename(rho = value)
spearP_processed_Clr <- reordered_spearP_Clr  %>% get_upper_tri() %>% reshape2::melt() %>% na.omit() %>% dplyr::rename(p = value)

# join the two data frames

sum(spearCor_processed_Clr$Var1 != spearP_processed_Clr$Var1)
sum(spearCor_processed_Clr$Var2 != spearP_processed_Clr$Var2)


spearRhoP_Clr <- cbind(spearCor_processed_Clr, p=spearP_processed_Clr$p) %>%
  # # remove self correlations
  filter(Var1 != Var2) %>% 
  # calculate the false discovery rate to adjust for multiple p values
  mutate(fdr = p.adjust(p, method = "BH"))


hist(spearRhoP_Clr$p)
```

```{r}
fdrThresh = 0.05

spearOkP_Clr <- spearRhoP_Clr %>% filter(fdr < fdrThresh) 

spearRhoPClr_plot <- spearRhoP_Clr %>% ggplot(aes(x = Var2, y = Var1, fill = rho)) + 
  geom_tile() + 
  scale_fill_gradient2() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  geom_point(data = spearOkP_Clr, shape = 1)

spearRhoPClr_plot
```


Trying to visualize just the significant results.
```{r}
library(igraph)
library(ggraph)
#unmodified plot
edges <- spearOkP_Clr

g <- graph_from_data_frame(edges, directed = FALSE)


degAll <- igraph::degree(g, v = igraph::V(g), mode = "all")
betAll <- igraph::betweenness(g, v = igraph::V(g), directed = FALSE) / (((igraph::vcount(g) - 1) * (igraph::vcount(g)-2)) / 2)
betAll.norm <- (betAll - min(betAll))/(max(betAll) - min(betAll))
rm(betAll)

dsAll <- igraph::similarity.dice(g, vids = igraph::V(g), mode = "all")

g <- igraph::set.vertex.attribute(g, "deg", index = igraph::V(g), value = degAll)
g <- igraph::set.vertex.attribute(g, "bet", index = igraph::V(g), value = betAll.norm)


deg <- igraph::degree(g)
bet <- igraph::betweenness(g, directed = F)

ggraph(g, layout = "kk") +
  geom_edge_link(aes(width = abs(rho), color = rho > 0), alpha = 0.8) +
  geom_node_point(size = log2(deg+1)+1) +
  geom_node_text(aes(label = name), repel = TRUE, size = 4) +
  scale_edge_width(range = c(0.5, 2.5)) +
  scale_edge_color_manual(values = c("TRUE" = "steelblue", "FALSE" = "firebrick")) +
  theme_void()




ggraph(g, layout = "kk") +
  geom_edge_link(
    aes(width = abs(rho), color = rho > 0),
    alpha = 0.4
  ) +
  geom_node_point(size = 2, color = "grey") +
  # geom_node_label(
  #   aes(label = name),
  #   fill = "white",
  #   color = "black",
  #   size = 3.5,
  #   repel = TRUE,
  #   label.padding = unit(0.15, "lines"),
  #   label.r = unit(0.1, "lines"),
  #   label.size = 0.25
  # ) +
  geom_node_label(
    aes(label = name, fill = deg),
    color = "white",
    repel = TRUE,
    size = 3.5,
    label.size = 0.25
  ) +
  scale_fill_viridis_c(option = "magma", end = 0.9) +
  scale_edge_width(range = c(0.5, 2.5)) +
  scale_edge_color_manual(
    values = c("TRUE" = "steelblue", "FALSE" = "firebrick")
  ) +
  theme_void()
```


```{r}
edges <- spearOkP_Clr

g <- graph_from_data_frame(edges, directed = FALSE)


degAll <- igraph::degree(g, v = igraph::V(g), mode = "all")
betAll <- igraph::betweenness(g, v = igraph::V(g), directed = FALSE) / (((igraph::vcount(g) - 1) * (igraph::vcount(g)-2)) / 2)
betAll.norm <- (betAll - min(betAll))/(max(betAll) - min(betAll))
rm(betAll)

dsAll <- igraph::similarity.dice(g, vids = igraph::V(g), mode = "all")

g <- igraph::set.vertex.attribute(g, "deg", index = igraph::V(g), value = degAll)
g <- igraph::set.vertex.attribute(g, "bet", index = igraph::V(g), value = betAll.norm)


deg <- igraph::degree(g)
bet <- igraph::betweenness(g, directed = F)

#renaming
#"CAG-56"
as.data.frame(tax_table(rare_r))[grepl("CAG-56", as.data.frame(tax_table(rare_r))$Genus), ]
as.data.frame(tax_table(rare_r))["f247d6391a08e1c052517539c487d0864a7d9ca9",]

V(g)$name[V(g)$name == "CAG-56"] <- "Lachnospiraceae CAG-56"

#"UBA1819"
as.data.frame(tax_table(rare_r))[grepl("UBA1819", as.data.frame(tax_table(rare_r))$Genus), ]
as.data.frame(tax_table(rare_r))["189345b67bd77229c8027462d9081974e7bd252c",]

V(g)$name[V(g)$name == "UBA1819"] <- "Ruminococcaceae UBA1819"

#"DTU089"
as.data.frame(tax_table(rare_r))[grepl("DTU089", as.data.frame(tax_table(rare_r))$Genus), ]
as.data.frame(tax_table(rare_r))["2dbd721f8a6f7678c32638f76877db3fbda70c51",]

V(g)$name[V(g)$name == "DTU089"] <- "Ruminococcaceae DTU089"

#"unclassified"

# reformat Genus text as follows:
# 1. Replace _ → space
# 2. If two parts:
# --First word (family like Ruminococcaceae) → NOT italic
# --Everything after → italic
# 3. If a trailing standalone number → keep number NOT italic
# 4. Otherwise (single genus like Prevotella, UBA1819) → fully italicunderscores with spaces + build italic labels

label_clean <- gsub("_", " ", V(g)$name)

make_label_expr <- function(x) {
  if (tolower(x) == "unclassified" ||
      x %in% c("Unclassified Christensenellaceae",
               "Unclassified Rhodospirillales")) {
    paste0("'", x, "'")
  } else {
    paste0("italic('", x, "')")
  }
}

V(g)$label_expr <- sapply(label_clean, make_label_expr)
#   # case 0: unclassified → NOT italic
#   if (tolower(x) == "unclassified") {
#     return(paste0("'", x, "'"))
#   }
#   
#   # case 1: trailing standalone number, e.g. Prevotella 7
#   if (grepl("^[A-Za-z]+ [0-9]+$", x)) {
#     return(sub("^(.*) ([0-9]+)$", "italic('\\1')~'\\2'", x))
#   }
#   
#   # case 2: Family XIII AD3011 group
#   if (grepl("^Family [A-Za-z0-9]+ ", x)) {
#     first <- sub("^(Family [A-Za-z0-9]+) (.*)$", "\\1", x)
#     rest  <- sub("^(Family [A-Za-z0-9]+) (.*)$", "\\2", x)
#     return(paste0("'", first, "'~italic('", rest, "')"))
#   }
#   
#   # case 3: standard family + group (one-word family)
#   if (grepl(" ", x)) {
#     first <- sub("^([^ ]+) (.*)$", "\\1", x)
#     rest  <- sub("^([^ ]+) (.*)$", "\\2", x)
#     return(paste0("'", first, "'~italic('", rest, "')"))
#   }
#   
#   # case 4: single word
#   paste0("italic('", x, "')")
# }

V(g)$label_expr <- sapply(label_clean, make_label_expr)

ggraph(g, layout = "kk") +
  geom_edge_link(
    aes(width = abs(rho), color = rho > 0),
    alpha = 0.4
  ) +
  geom_node_point(size = 2, color = "grey") +
  geom_node_label(
    aes(label = label_expr, fill = deg),
    parse = TRUE,
    color = "white",
    repel = TRUE,
    max.overlaps = Inf,
    size = 3.5,
    label.size = 0.25
  ) +
  scale_fill_viridis_c(
    name = "Degree",              # rename deg
    option = "magma",
    end = 0.9
  ) +
  scale_edge_width(
    name = "Magnitude",           # rename abs(rho)
    range = c(0.5, 2.5)
  ) +
  scale_edge_color_manual(
    name = "Correlation",         # rename rho > 0
    values = c("TRUE" = "steelblue", "FALSE" = "firebrick"),
    labels = c("FALSE" = "Negative", "TRUE" = "Positive")  # relabel values
  ) +
  theme_void() +
  theme(
    legend.title = element_text(size = 10),
    legend.text  = element_text(size = 9)
  )
```
