---
title: "endo_vaginal_v9_final"
author: "Cameron A. Gorzney, Stephen Johnson, Raunak Vijayakar, Michelle Bland"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries

```{r}
###r version 4.3.2

# # Install specific Bioconductor version
# install.packages("BiocManager")
# 
# BiocManager::install(version = "3.17")
# 
# # Then install packages
# BiocManager::install(c("phyloseq", "microbiome"))
```


```{r}
# packageVersion("")
library(vegan) #remotes::install_version("vegan", version = "2.6-10")
library(ggrepel)
library(readxl) #remotes::install_version("readxl", version = "1.4.5")  
library(phyloseq) #see above
library(ggplot2) #remotes::install_version("ggplot2", version = "4.0.0")  
library(hrbrthemes) #remotes::install_version("hrbrthemes", version = "0.8.7")  
library(ggsci) #remotes::install_version("ggsci", version = "3.2.0")  
library(dplyr) #remotes::install_version("dplyr", version = "1.1.4")  
library(microViz) #see below #needed to install "ggiraph" and "ComplexHeatmap"
# install.packages(
#   "microViz",
#   repos = c(davidbarnett = "https://david-barnett.r-universe.dev", getOption("repos"))
# )
library(microbiome) #see above
library(MicrobiomeStat) #remotes::install_version("MicrobiomeStat", version = "1.0")  
library(GUniFrac) #remotes::install_version("GUniFrac", version = "1.8")
library(picante) #remotes::install_version("picante", version = "1.8.2")
library(ggtext) #remotes::install_version("ggtext", version = "2.6-10")
library(tidyr)
library(purrr)
```

## Process samples

```{r}

setwd()

phylo = readRDS("data/Phylo_public.rds")
meta = samdat_tbl(phylo)

#look at metadata
summarize_phyloseq(phylo)

table(meta$Source) #I thought this was a binary variable? We'll use only vaginal and rectal
table(meta$Plate, meta$case_control) #maybe include plate as covariate in models?

```

Add sample library size.

```{r}
depth = data.frame(lib_size = sample_sums(phylo))
depth$SampleID = rownames(depth)
depth$SampleID = sub('.', '', depth$SampleID)

phylo = ps_join(x = phylo, y = depth, by = "SampleID", type = "inner", match_sample_names = NULL)

sample_data(phylo)$subject = as.factor(sample_data(phylo)$subject)
```

```{r}
unique(meta$Source)
unique(meta$Amp.Status)

phylo_r = phylo |> ps_filter(Source == "vagina") 
temp = sample_data(phylo_r)
unique(temp$Subject.ID)

colSums(table(temp$Subject.ID, temp$case_control) != 0) #check sample size

phylo_r = phylo_r |> ps_filter(Amp.Status == "Amp")
meta_r = samdat_tbl(phylo_r)
unique(meta_r$Subject.ID)

colSums(table(meta_r$Subject.ID, meta_r$case_control) != 0) #check sample size

#list the subjects who were lost after filtering for amplification
unique(temp$Subject.ID[!(temp$Subject.ID %in% meta_r$Subject.ID)])

```
Merge and filter.

```{r}
merged_r = merge_samples(phylo_r, "subject", fun = sum)
# We use the sum function here instead of mean to get correct lib_size.
# The other numeric metadata are completely wrecked, as they would be with any function.

# Keep subject IDs from the merged phyloseq object
sample_data(merged_r)$Subject.ID = sample_names(merged_r)

# Add corrected patient demographics metadata
newdata = read_xlsx(
 "data/Newdata_endo_pain.xlsx",col_names = T, sheet =1)


splits = strsplit(newdata$`Study ID`, split = "-")
splits = lapply(splits, head, 2)
newdata$Subject.ID = sapply(splits, paste, collapse = "")


# Keep all clinical variables from the corrected demographics sheet
newdata = newdata |> 
  dplyr::rename(
    case_control_new = `case vs control`,
    paratubal_cysts_new = `Paratubal Cysts`,
    Gravidity_new = Graviditity,
    Age_new = `Age at surgery`,
    parity_status_new= `Parity status`,
    parity_new = Parity,
    BMI_new = BMI,
    stage_new = stage,
    Severity_new = Severity,
    Endosalpingiosis_new = Endosalpingiosis,
    Adenomyosis_new = Adenomyosis,
    Fibroids_new = Fibroids,
    pain_1st_level_new = pain_1st_level,
    abnormal_bleeding_new = Abnormal_bleeding
    
  ) |>
  dplyr::select(
    Subject.ID,
    case_control_new,
    BMI_new,
    Age_new,
    Gravidity_new,
    stage_new,
    Severity_new,
    parity_status_new,
    Endosalpingiosis_new,
    Adenomyosis_new,
    Fibroids_new,
    paratubal_cysts_new,
    parity_new,
    pain_1st_level_new,
    abnormal_bleeding_new,
    Ethnicity,
    Race
  ) |>
  dplyr::distinct(Subject.ID, .keep_all = TRUE)

merged_r = ps_join(merged_r, newdata, by = "Subject.ID", type = "left")


# Number of unique participants with BOTH microbiome + pain data
n_distinct(sample_data(merged_r)$Subject.ID)

# Check that corrected demographics are still present after join
merged_meta_check = samdat_tbl(merged_r)

colnames(merged_meta_check)

table(merged_meta_check$case_control_new, useNA = "ifany")
summary(merged_meta_check$BMI_new)
summary(merged_meta_check$Age_new)
summary(merged_meta_check$parity_new)
table(merged_meta_check$Severity_new, useNA = "ifany")


merged_r <- merged_r %>%
  ps_filter(lib_size > 2000) %>%
  tax_filter(min_prevalence = 2)

merged_meta_r = samdat_tbl(merged_r)

merged_meta_r$case_control_new = as.factor(merged_meta_r$case_control_new)
merged_meta_r$case_control_new = relevel(merged_meta_r$case_control_new, ref = "control")
merged_meta_r$Fibroids_new = as.factor(merged_meta_r$Fibroids_new)
merged_meta_r$paratubal_cysts_new = as.factor(merged_meta_r$paratubal_cysts_new)
merged_meta_r$Endosalpingiosis_new = as.factor(merged_meta_r$Endosalpingiosis_new)
merged_meta_r$Adenomyosis_new = as.factor(merged_meta_r$Adenomyosis_new)
merged_meta_r$abnormal_bleeding_new = as.factor(merged_meta_r$abnormal_bleeding_new)


#check for library size confounding
summary(merged_meta_r$lib_size)
hist(merged_meta_r$lib_size)
boxplot(merged_meta_r$lib_size ~ merged_meta_r$case_control_new, notch = F)
wilcox.test(merged_meta_r$lib_size ~ merged_meta_r$case_control_new)

#rarefy
set.seed(123)
rare_r = rarefy_even_depth(merged_r, replace = F) #rarefy to min depth or 2k, whichever greater

```
## Checking samples that are matched  
```{r}
meta_amp <- samdat_tbl(
  phylo |> ps_filter(Amp.Status == "Amp")
)

matched_subjects_amp <- meta_amp %>%
  dplyr::filter(Source %in% c("vagina", "rectum")) %>%
  dplyr::group_by(Subject.ID) %>%
  dplyr::summarise(
    n_sources = dplyr::n_distinct(Source)
  ) %>%
  dplyr::filter(n_sources == 2)

nrow(matched_subjects_amp)

matched_subjects_amp$Subject.ID

matched_case_control <- meta_amp %>%
  dplyr::filter(Subject.ID %in% matched_subjects_amp$Subject.ID) %>%
  dplyr::distinct(Subject.ID, case_control)

table(matched_case_control$case_control)

```

### Checking other diagnoses against case/control status
```{r}
categorical_pvals <- tibble::tibble(
  Variable = c(
    "Adenomyosis",
    "Endosalpingiosis",
    "Fibroids",
    "Paratubal Cysts",
    "Abnormal Bleeding",
    "Race",
    "Ethnicity"
  ),
  p_value = c(
    fisher.test(table(merged_meta_r$Adenomyosis_new,
                      merged_meta_r$case_control_new))$p.value,

    fisher.test(table(merged_meta_r$Endosalpingiosis_new,
                      merged_meta_r$case_control_new))$p.value,

    fisher.test(table(merged_meta_r$Fibroids_new,
                      merged_meta_r$case_control_new))$p.value,

    fisher.test(table(merged_meta_r$paratubal_cysts_new,
                      merged_meta_r$case_control_new))$p.value,

    fisher.test(table(merged_meta_r$abnormal_bleeding_new,
                      merged_meta_r$case_control_new))$p.value,

    fisher.test(table(merged_meta_r$Race,
                      merged_meta_r$case_control_new))$p.value,

    fisher.test(table(merged_meta_r$Ethnicity,
                      merged_meta_r$case_control_new))$p.value
  )
) %>%
  dplyr::mutate(
    FDR = p.adjust(p_value, method = "BH")
  )

categorical_pvals

### Continous Summary
continuous_summary <- merged_meta_r %>%
  dplyr::filter(
    !is.na(case_control_new),
    case_control_new %in% c("case", "control")
  ) %>%
  dplyr::select(
    case_control_new,
    Age_new,
    BMI_new,
    Gravidity_new,
    parity_new
  ) %>%
  tidyr::pivot_longer(
    cols = c(Age_new, BMI_new, Gravidity_new, parity_new),
    names_to = "Variable",
    values_to = "Value"
  ) %>%
  dplyr::group_by(Variable, case_control_new) %>%
  dplyr::summarise(
    mean = mean(Value, na.rm = TRUE),
    sd = sd(Value, na.rm = TRUE),
    mean_sd = paste0(round(mean, 2), " ± ", round(sd, 2)),
    n = sum(!is.na(Value)),
    .groups = "drop"
  )

continuous_summary

continuous_pvals <- merged_meta_r %>%
  dplyr::filter(
    !is.na(case_control_new),
    case_control_new %in% c("case", "control")
  ) %>%
  dplyr::select(
    case_control_new,
    Age_new,
    BMI_new,
    Gravidity_new,
    parity_new
  ) %>%
  tidyr::pivot_longer(
    cols = c(Age_new, BMI_new, Gravidity_new, parity_new),
    names_to = "Variable",
    values_to = "Value"
  ) %>%
  dplyr::group_by(Variable) %>%
  dplyr::summarise(
    p_value = wilcox.test(Value ~ case_control_new)$p.value,
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    FDR = p.adjust(p_value, method = "BH")
  )

continuous_pvals

```

### New bar
```{r}
## Analyze rectal samples by case/control

rare_r <- rare_r %>% tax_fix(unknowns = c("unclassified"))

sample_data(rare_r)$Subject.ID <- sample_names(rare_r)

sample_data(rare_r)$case_control_new <- merged_meta_r$case_control_new[
  match(sample_data(rare_r)$Subject.ID, merged_meta_r$Subject.ID)
]

table(sample_data(rare_r)$case_control_new, useNA = "ifany")

rare_r_case <- rare_r %>%
  ps_filter(
    !is.na(case_control_new),
    tolower(as.character(case_control_new)) %in% c("case", "control")
  ) %>%
  tax_filter(min_prevalence = 1)

case_raw <- tolower(as.character(sample_data(rare_r_case)$case_control_new))

sample_data(rare_r_case)$case_control_plot <- dplyr::case_when(
  case_raw == "control" ~ "Control",
  case_raw == "case" ~ "Case",
  TRUE ~ NA_character_
)

case_counts <- table(sample_data(rare_r_case)$case_control_plot)

sample_data(rare_r_case)$case_control_plot <- paste0(
  sample_data(rare_r_case)$case_control_plot,
  " (n = ",
  case_counts[sample_data(rare_r_case)$case_control_plot],
  ")"
)

sample_data(rare_r_case)$case_control_plot <- factor(
  sample_data(rare_r_case)$case_control_plot,
  levels = c(
    paste0("Control (n = ", case_counts["Control"], ")"),
    paste0("Case (n = ", case_counts["Case"], ")")
  )
)

table(sample_data(rare_r_case)$case_control_plot, useNA = "ifany")

tax_df <- as.data.frame(tax_table(rare_r_case), stringsAsFactors = FALSE)

tax_df$Genus[tax_df$Genus == "Lachnospiraceae Family"] <- 
  "Lachnospiraceae Unclassified"

tax_df$Genus[tax_df$Genus == "Ruminococcaceae Family"] <- 
  "Ruminococcaceae Unclassified"

tax_df$Genus[tax_df$Genus == "UBA1819"] <- 
  "Ruminococcaceae UBA1819"

tax_table(rare_r_case) <- as.matrix(tax_df)

bar_case <- rare_r_case %>% 
  comp_barplot(
    tax_level = "Genus",
    label = "Subject.ID",
    n_taxa = 20,
    taxon_renamer = function(x) {
      
      x_clean <- stringr::str_replace_all(x, "_", " ")
      
      dplyr::case_when(
        x_clean == "Lachnospiraceae Unclassified" ~
          "<i>Lachnospiraceae</i> Unclassified",
        
        x_clean == "Ruminococcaceae Unclassified" ~
          "<i>Ruminococcaceae</i> Unclassified",
        
        x_clean == "Other genera" ~ x_clean,
        
        TRUE ~ paste0("<i>", x_clean, "</i>")
      )
    },
    other_name = "Other genera",
    merge_other = TRUE,
    bar_width = 1,
    bar_outline_colour = "grey5",
    seriate_method = "OLO_ward",
    facet_by = "case_control_plot",
    sample_order = sample_names(rare_r_case)
  ) + 
  coord_flip() + 
  guides(fill = guide_legend(ncol = 1)) +
  labs(
    y = "Relative Abundance",
    x = NULL,
    fill = "Genus"
  ) +
  theme(
    strip.text = element_text(color = "black", size = 20, face = "bold"),
    axis.text.x = element_text(color = "black", size = 20, margin = margin(t = 6)), 
    axis.text.y = element_text(color = "black", size = 20, margin = margin(r = 4)), 
    axis.title.x = element_text(size = 20, face = "bold", margin = margin(t = 8)),
    axis.title.y = element_text(size = 20, face = "bold", margin = margin(t = 8)), 
    axis.ticks.x = element_line(linewidth = 0.8),
    axis.ticks.length.x = unit(0.18, "cm"),
    legend.title = element_text(size = 15, face = "bold"),
    legend.text = ggtext::element_markdown(size = 13, lineheight = 0.85),
    legend.key.size = unit(0.35, "cm"),
    legend.spacing.y = unit(0.05, "cm"),
    legend.margin = margin(0, 0, 0, 0),
    panel.background = element_rect(fill = "transparent", colour = NA),
    panel.spacing.x = unit(0.75, "cm"),
    plot.background = element_rect(fill = "transparent", colour = NA),
    legend.background = element_rect(fill = "transparent", colour = NA)
  )

bar_case


library(cowplot)

legend_case <- get_legend(
  bar_case +
    theme(
      legend.background = element_rect(fill = "transparent", colour = NA),
      legend.box.background = element_rect(fill = "transparent", colour = NA),
      legend.key = element_rect(fill = "transparent", colour = NA),
      panel.background = element_rect(fill = "transparent", colour = NA),
      plot.background = element_rect(fill = "transparent", colour = NA)
    )
)

bar_case_nolegend <- bar_case +
  theme(
    legend.position = "none",
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA)
  )

bar_case_nolegend
legend_case

```

## Save bar plot
```{r}
ggsave(
  "vaginal_ra_nolegend.pdf",
  plot = bar_case_nolegend,
  path = fig_dir,
  width = 14,
  height = 10,
  bg = "transparent"
)
ggsave(
  "legend_case.pdf",
  plot = legend_case,
  path = fig_dir,
  width = 3,
  height = 5,
  bg = "transparent"
)

```


### Extract Bar and Legend Separately for Figures
```{r}
library(cowplot)

legend <- get_legend(
  bar +
    theme(
    legend.background = element_rect(
      fill = "transparent",
      colour = NA
    ),
    legend.box.background = element_rect(
      fill = "transparent",
      colour = NA
    ),
    legend.key = element_rect(
      fill = "transparent",
      colour = NA
    ),
    panel.background = element_rect(
      fill = "transparent",
      colour = NA
    ),
    plot.background = element_rect(
      fill = "transparent",
      colour = NA
    )
))


bar_nolegend <- bar +
  theme(
    legend.position = "none",
    panel.background = element_rect(
      fill = "transparent",
      colour = NA
    ),
    plot.background = element_rect(
      fill = "transparent",
      colour = NA
    )
  )

bar_nolegend

ggsave(
  "vaginal_ra_nolegend.pdf",
  plot = bar_nolegend,
  path = fig_dir,
  width = 14,
  height = 10,
  bg = "transparent"
)
ggsave(
  "vaginal.pdf",
  plot = legend,
  path = fig_dir,
  width = 3,
  height = 5,
  bg = "transparent"
)

ggsave(
  filename = "vaginal_ra.png",
  plot = bar,
  path = fig_dir,
  width = 16,
  height = 10,
  units = "in",
  dpi = 600,
  bg = "transparent"
)

ggsave(
  filename = "vaginal_ra.pdf",
  plot = bar,
  path = fig_dir,
  width = 16,
  height = 10,
  units = "in",
  device = "pdf",
  bg = "transparent"
)

```


### Alpha Diversity 

```{r}
set.seed(123)

metric_labels <- c(
  Observed = "Observed ASVs",
  Chao1 = "Chao1 Richness",
  Shannon = "Shannon Diversity Index",
  InvSimpson = "Inverse Simpson Index"
)

plot_richness(rare_r, x="case_control_new", measures=c("Observed", "Chao1", "Shannon", "InvSimpson")) + geom_boxplot()

rich_r <- estimate_richness(
  rare_r,
  measures = c("Observed", "Chao1", "Shannon", "InvSimpson")
)

rich_r$case_control_new <- sample_data(rare_r)$case_control_new

rich_r <- rich_r %>%
  mutate(
    Subject.ID = sample_data(rare_r)$Subject.ID,
    case_control_new = as.character(sample_data(rare_r)$case_control_new),
    case_control_new = dplyr::case_when(
      stringr::str_detect(stringr::str_to_lower(case_control_new), "control") ~ "Controls\n(n = 9)",
      stringr::str_detect(stringr::str_to_lower(case_control_new), "case") ~ "Cases\n(n = 20)",
      TRUE ~ NA_character_
    ),
    case_control_new = factor(
      case_control_new,
      levels = c("Controls\n(n = 9)", "Cases\n(n = 20)")
    )
  )

table(rich_r$case_control_new, useNA = "ifany")

rich_r <- rich_r %>%
  left_join(
    merged_meta_r %>%
      dplyr::select(
        Subject.ID,
        parity_new,
        parity_status_new,
        Gravidity_new,
        BMI_new,
        Age_new
      ),
    by = "Subject.ID"
  )
table(rich_r$case_control_new)

alpha_results <- tibble(
  Metric = c("Observed", "Chao1", "Shannon", "InvSimpson"),
  Pvalue = c(
    wilcox.test(Observed ~ case_control_new, data = rich_r)$p.value,
    wilcox.test(Chao1 ~ case_control_new, data = rich_r)$p.value,
    wilcox.test(Shannon ~ case_control_new, data = rich_r)$p.value,
    wilcox.test(InvSimpson ~ case_control_new, data = rich_r)$p.value
  )
) %>%
  mutate(
    FDR = p.adjust(Pvalue, method = "fdr"),
    p_label = paste0("Wilcoxon p = ", signif(Pvalue, 3))
  )

alpha_results

plot_alpha_metric <- function(metric_name) {
  
  p_value <- alpha_results %>%
    filter(Metric == metric_name) %>%
    pull(Pvalue)
  
  p_lab <- bquote("Wilcoxon " * italic(p) == .(signif(p_value, 3)))
  # p_lab <- alpha_results %>%
  #   dplyr::filter(Metric == metric_name) %>%
  #   dplyr::pull(p_label)
  
  ggplot(
    rich_r,
    aes(
      x = case_control_new,
      y = .data[[metric_name]],
      fill = case_control_new
    )
  ) +
    geom_boxplot(
      outlier.shape = NA,
      alpha = 0.75,
      color = "black"
    ) +
    geom_jitter(
      width = 0.12,
      size = 3,
      alpha = 0.85,
      color = "black"
    ) +
    scale_fill_manual(
      values = c(
        "Controls\n(n = 9)" = "#5B84B1",
        "Cases\n(n = 20)" = "#D77A9B"
      )
    ) +
    theme_classic(base_size = 16) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(angle = 45, hjust = 1, size = 18, face = "bold"),
      axis.text.y = element_text(size = 18),
      axis.title.x = element_text(size = 18),
      axis.title.y = element_text(size = 18, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 24, face = "bold"),
      panel.background = element_rect(fill = "transparent", colour = NA),
      plot.background = element_rect(fill = "transparent", colour = NA)
    ) +
    labs(
      x = NULL,
      y = metric_labels[[metric_name]],
      subtitle = p_lab
      #title = paste0(metric_name, " by case/control status")
    )
}

p_observed <- plot_alpha_metric("Observed")
p_chao1 <- plot_alpha_metric("Chao1")
p_shannon <- plot_alpha_metric("Shannon")
p_invsimpson <- plot_alpha_metric("InvSimpson")

p_observed
p_chao1
p_shannon
p_invsimpson
```

## Save alpha diversity figures
```{r}
ggsave(
  filename = "Observed_alpha_diversity.png",
  plot = p_observed,
  path = fig_dir,
  width = 6,
  height = 5.5,
  units = "in",
  dpi = 600,
  bg = "transparent"
)

ggsave(
  filename = "Observed_alpha_diversity.pdf",
  plot = p_observed,
  path = fig_dir,
  width = 6,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)

ggsave(
  filename = "Chao1_alpha_diversity.png",
  plot = p_chao1,
  path = fig_dir,
  width = 6,
  height = 5.5,
  units = "in",
  dpi = 600,
  bg = "transparent"
)

ggsave(
  filename = "Chao1_alpha_diversity.pdf",
  plot = p_chao1,
  path = fig_dir,
  width = 6,
  height = 5.5,
  units = "in",
  dpi = 600,
  bg = "transparent"
)

ggsave(
  filename = "Shannon_alpha_diversity.png",
  plot = p_shannon,
  path = fig_dir,
  width = 6,
  height = 5.5,
  units = "in",
  dpi = 600,
  bg = "transparent"
)

ggsave(
  filename = "Shannon_alpha_diversity.pdf",
  plot = p_shannon,
  path = fig_dir,
  width = 6,
  height = 5.5,
  units = "in",
  dpi = 600,
  bg = "transparent"
)
ggsave(
  filename = "InvSimpson_alpha_diversity.png",
  plot = p_invsimpson,
  path = fig_dir,
  width = 6,
  height = 5.5,
  units = "in",
  dpi = 600,
  bg = "transparent"
)

ggsave(
  filename = "InvSimpson_alpha_diversity.pdf",
  plot = p_invsimpson,
  path = fig_dir,
  width = 6,
  height = 5.5,
  units = "in",
  dpi = 600,
  bg = "transparent"
)
```

Phylogenetic diversity from `picante`.

```{r}
set.seed(123)

pd_result <- pd(
  otu_table(rare_r),
  phy_tree(rare_r),
  include.root = TRUE
)

pd_df <- pd_result %>%
  as.data.frame()

pd_df$Subject.ID <- rownames(pd_df)

pd_df$case_control_new <- sample_data(rare_r)$case_control_new

pd_df <- pd_df %>%
  left_join(
    merged_meta_r %>%
      dplyr::select(
        Subject.ID,
        parity_new,
        parity_status_new,
        Gravidity_new,
        BMI_new,
        Age_new
      ),
    by = "Subject.ID"
  )

pd_df <- pd_df %>%
  mutate(
    case_control_new = factor(
      case_control_new,
      levels = c("control", "case"),
      labels = c("Controls\n(n = 9)", "Cases\n(n = 20)")
    )
  )

pd_test <- wilcox.test(
  PD ~ case_control_new,
  data = pd_df
)

pd_test

pd_label <- bquote("Wilcoxon " * italic(p) == .(signif(pd_test$p.value, 3)))

pd_plot <- ggplot(
  pd_df,
  aes(
    x = case_control_new,
    y = PD,
    fill = case_control_new
  )
) +
  geom_boxplot(
    outlier.shape = NA,
    alpha = 0.75,
    color = "black"
  ) +
  geom_jitter(
    width = 0.12,
    size = 3,
    alpha = 0.85,
    color = "black"
  ) +
  scale_fill_manual(
    values = c(
      "Controls\n(n = 9)" = "#5B84B1",
      "Cases\n(n = 20)" = "#D77A9B"
    )
  ) +
  theme_classic(base_size = 16) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 18, face = "bold"),
    axis.text.y = element_text(size = 18),
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 24),
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA)
  ) +
  labs(
    x = NULL,
    y = "Faith's Phylogenetic Diversity",
    #title = "Phylogenetic diversity by case/control status",
    subtitle = pd_label
  )

pd_plot
```

# Save Figures

```{r}
ggsave(
  filename = "Faith_PD.png",
  plot = pd_plot,
  path = fig_dir,
  width = 6,
  height = 5.5,
  units = "in",
  dpi = 600,
  bg = "transparent"
)

ggsave(
  filename = "Faith_PD.pdf",
  plot = pd_plot,
  path = fig_dir,
  width = 6,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)
```

### Beta diversity by case/control

```{r}
set.seed(123)

# Match metadata to rare_r sample order
merged_meta_r <- merged_meta_r %>%
  dplyr::filter(Subject.ID %in% sample_names(rare_r)) %>%
  dplyr::arrange(match(Subject.ID, sample_names(rare_r))) %>%
  dplyr::mutate(
    case_control_new = factor(
      case_control_new,
      levels = c("control", "case"),
      labels = c("Controls (n = 9)", "Cases (n = 20)")
    ),
    parity_status_new = factor(parity_status_new)
  )

stopifnot(all(merged_meta_r$Subject.ID == sample_names(rare_r)))

# ----------------------------------------------------------
# Build distance matrices
# ----------------------------------------------------------

tree_r <- phy_tree(rare_r)

otu_r <- as(otu_table(rare_r), "matrix")

if (taxa_are_rows(rare_r)) {
  otu_r <- t(otu_r)
}

otu_r <- otu_r[merged_meta_r$Subject.ID, ]

unifracs <- GUniFrac(
  otu.tab = otu_r,
  tree = tree_r,
  alpha = c(0, 0.5, 1)
)$unifracs

uni_dist <- as.dist(unifracs[, , "d_UW"])
guni_dist <- as.dist(unifracs[, , "d_0.5"])
wuni_dist <- as.dist(unifracs[, , "d_1"])

jac_dist <- vegdist(
  otu_r,
  method = "jaccard",
  binary = TRUE
)

bray_dist <- vegdist(
  otu_r,
  method = "bray"
)

### Prepare pseudocounts for aitchison
# Genus-level non-rarefied counts
merged_r_genus <- tax_glom(merged_r, taxrank = "Genus")

genus_counts <- as.data.frame(otu_table(merged_r_genus))

# Make sure rows = subjects and columns = taxa
if (taxa_are_rows(merged_r_genus)) {
  genus_counts <- as.data.frame(t(genus_counts))
}

# Add clean genus names as column names
genera <- as.data.frame(tax_table(merged_r_genus))
genus_names <- as.character(genera$Genus)
genus_names[is.na(genus_names) | genus_names == ""] <- "unclassified"
genus_names <- make.unique(genus_names)

colnames(genus_counts) <- genus_names

# CLR transform
genus_counts_pseudo <- genus_counts + 1
clr_mat <- compositions::clr(genus_counts_pseudo)

# Aitchison PCA
pca1 <- prcomp(clr_mat, center = TRUE, scale. = FALSE)
summary(pca1)

scores <- as.data.frame(pca1$x)
scores$Subject.ID <- rownames(scores)

# Aitchison distance
aitch <- vegan::vegdist(clr_mat, method = "euclidean")

# ----------------------------------------------------------
# PERMANOVA
# ----------------------------------------------------------

uni_test <- adonis3(uni_dist ~ case_control_new, data = merged_meta_r)
guni_test <- adonis3(guni_dist ~ case_control_new, data = merged_meta_r)
jac_test <- adonis3(jac_dist ~ case_control_new, data = merged_meta_r)
wuni_test <- adonis3(wuni_dist ~ case_control_new, data = merged_meta_r)
bray_test <- adonis3(bray_dist ~ case_control_new, data = merged_meta_r)
aitch_test <- adonis3(aitch ~ case_control_new, data = merged_meta_r)


#### Beta Diversity Results  for case-control, run this

beta_results <- data.frame(
  Metric = c(
    "UniFrac",
    "Generalized UniFrac",
    "Jaccard",
    "Weighted UniFrac",
    "Bray-Curtis",
    "Aitchison Distance"
  ),
  R2 = c(
    uni_test$aov.tab["case_control_new", "R2"],
    guni_test$aov.tab["case_control_new", "R2"],
    jac_test$aov.tab["case_control_new", "R2"],
    wuni_test$aov.tab["case_control_new", "R2"],
    bray_test$aov.tab["case_control_new", "R2"],
    aitch_test$aov.tab["case_control_new", "R2"]
  ),
  Pvalue = c(
    uni_test$aov.tab["case_control_new", "Pr(>F)"],
    guni_test$aov.tab["case_control_new", "Pr(>F)"],
    jac_test$aov.tab["case_control_new", "Pr(>F)"],
    wuni_test$aov.tab["case_control_new", "Pr(>F)"],
    bray_test$aov.tab["case_control_new", "Pr(>F)"],
    aitch_test$aov.tab["case_control_new", "Pr(>F)"]
  )
) %>%
  dplyr::mutate(
    FDR = p.adjust(Pvalue, method = "fdr")
  )

beta_results

# ----------------------------------------------------------
# Plot settings
# ----------------------------------------------------------

case_cols <- c(
  "Controls (n = 9)" = "#5B84B1",
  "Cases (n = 20)" = "#D77A9B"
)

get_p <- function(metric_name) {
  signif(beta_results$Pvalue[beta_results$Metric == metric_name], 3)
}

get_adj_p <- function(metric_name) {
  signif(beta_results$FDR[beta_results$Metric == metric_name], 3)
}

# ----------------------------------------------------------
# PCoA plots from phyloseq-supported distances
# ----------------------------------------------------------
sample_data(rare_r)$case_control_new <- factor(
  sample_data(rare_r)$case_control_new,
  levels = c("control", "case"),
  labels = c(
    "Controls (n = 9)",
    "Cases (n = 20)"
  )
)

ord_uni <- ordinate(rare_r, "PCoA", "unifrac")

p_uni <- plot_ordination(rare_r, ord_uni, color = "case_control_new") +
  stat_ellipse(
    aes(group = case_control_new),
    linewidth = 0.5,
    show.legend = FALSE
  ) +
  geom_point(size = 3.5) +
  scale_color_manual(values = case_cols) +
  theme_classic(base_size = 16) +
  theme(
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 20, face = "bold"),
    axis.title.y = element_text(size = 20, face = "bold"),
    plot.title = element_text(size = 24, hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(size = 22, hjust = 0.5),
    legend.title = element_blank(),
    legend.text = element_text(size = 18),
    legend.margin = margin(0, 0, 0, 0),
    
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank()
  ) +
  labs(
    #title = "PCoA using UniFrac",
    x = paste0("PCoA1 [", round(ord_uni$values$Relative_eig[1] * 100, 1), "%]"),
    y = paste0("PCoA2 [", round(ord_uni$values$Relative_eig[2] * 100, 1), "%]"),
    title = bquote("UniFrac"),
    subtitle = bquote("Adjusted " * italic(p) == .(get_adj_p("UniFrac")))#,
    #color = "Case/control"
  )

ord_jac <- ordinate(rare_r, "PCoA", "jaccard")

p_jac <- plot_ordination(rare_r, ord_jac, color = "case_control_new") +
  stat_ellipse(
    aes(group = case_control_new),
    linewidth = 0.5,
    show.legend = FALSE
  ) +
  geom_point(size = 3.5) +
  scale_color_manual(values = case_cols) +
  theme_classic(base_size = 16) +
  theme(
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 20, face = "bold"),
    axis.title.y = element_text(size = 20, face = "bold"),
    plot.title = element_text(size = 24, hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(size = 22, hjust = 0.5),
    legend.title = element_blank(),
    legend.text = element_text(size = 18),
    legend.margin = margin(0, 0, 0, 0),
    
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank()
  ) +
  labs(
    #title = "PCoA using Jaccard",
    x = paste0("PCoA1 [", round(ord_jac$values$Relative_eig[1] * 100, 1), "%]"),
    y = paste0("PCoA2 [", round(ord_jac$values$Relative_eig[2] * 100, 1), "%]"),
    title = bquote("Jaccard"),
    subtitle = bquote("Adjusted " * italic(p) == .(get_adj_p("Jaccard")))#,
    #color = "Case/control"
  )

ord_wuni <- ordinate(rare_r, "PCoA", "wunifrac")

p_wuni <- plot_ordination(rare_r, ord_wuni, color = "case_control_new") +
  stat_ellipse(
    aes(group = case_control_new),
    linewidth = 0.5,
    show.legend = FALSE
  ) +
  geom_point(size = 3.5) +
  scale_color_manual(values = case_cols) +
  theme_classic(base_size = 16) +
  theme(
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 20, face = "bold"),
    axis.title.y = element_text(size = 20, face = "bold"),
    plot.title = element_text(size = 24, hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(size = 22, hjust = 0.5),
    legend.title = element_blank(),
    legend.text = element_text(size = 18),
    legend.margin = margin(0, 0, 0, 0),
    
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank()
  ) +
  labs(
    #title = "PCoA using Weighted UniFrac",
    x = paste0("PCoA1 [", round(ord_wuni$values$Relative_eig[1] * 100, 1), "%]"),
    y = paste0("PCoA2 [", round(ord_wuni$values$Relative_eig[2] * 100, 1), "%]"),
    title = bquote("Weighted UniFrac"),
    subtitle = bquote("Adjusted " * italic(p) == .(get_adj_p("Weighted UniFrac")))#,
    #color = "Case/control"
  )

ord_bray <- ordinate(rare_r, "PCoA", "bray")

p_bray <- plot_ordination(rare_r, ord_bray, color = "case_control_new") +
  stat_ellipse(
    aes(group = case_control_new),
    linewidth = 0.5,
    show.legend = FALSE
  ) +
  geom_point(size = 3.5) +
  scale_color_manual(values = case_cols) +
  theme_classic(base_size = 16) +
  theme(
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 20, face = "bold"),
    axis.title.y = element_text(size = 20, face = "bold"),
    plot.title = element_text(size = 24, hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(size = 22, hjust = 0.5),
    legend.title = element_blank(),
    legend.text = element_text(size = 18),
    legend.margin = margin(0, 0, 0, 0),
    
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank()
  ) +
  labs(
    #title = "PCoA using Bray-Curtis",
    x = paste0("PCoA1 [", round(ord_bray$values$Relative_eig[1] * 100, 1), "%]"),
    y = paste0("PCoA2 [", round(ord_bray$values$Relative_eig[2] * 100, 1), "%]"),
    title = bquote ("Bray-Curtis"),
    subtitle = bquote("Adjusted " * italic(p) == .(get_adj_p("Bray-Curtis")))#,
    #color = "Case/control"
  )

p_bray
# PCA scores
scores <- as.data.frame(pca1$x)
scores$Subject.ID <- rownames(scores)

# Verify IDs match metadata
setequal(scores$Subject.ID, merged_meta_r$Subject.ID)


# Add original case/control metadata
scores <- scores %>%
  dplyr::left_join(
    merged_meta_r %>%
      dplyr::select(Subject.ID, case_control_new),
    by = "Subject.ID"
  ) %>%
  dplyr::mutate(
    case_control_plot = dplyr::case_when(
      grepl("Control", as.character(case_control_new), ignore.case = TRUE) ~ "Controls (n = 9)",
      grepl("Case", as.character(case_control_new), ignore.case = TRUE) ~ "Cases (n = 20)",
      TRUE ~ NA_character_
    )
  )

table(scores$case_control_plot, useNA = "always")
# Axis labels
var_exp <- (pca1$sdev^2) / sum(pca1$sdev^2)

xlab <- paste0("PC1 [", round(100 * var_exp[1], 1), "%]")
ylab <- paste0("PC2 [", round(100 * var_exp[2], 1), "%]")

#Add Aitchison axes and clusters directly to merged_meta_r
merged_meta_r <- merged_meta_r %>%
dplyr::select(
     -any_of(c(
       "PC1",
       "PC2"
     ))
   ) %>%
   dplyr::left_join(
     scores %>%
       dplyr::select(
         Subject.ID,
         PC1,
         PC2,
       ),
     by = "Subject.ID"
   )

# Aitchison PCA plot
p_aitch <- ggplot(
   merged_meta_r,
   aes(
     x = PC1,
     y = PC2,
     color = case_control_new
   )
 ) +
 
   stat_ellipse(
     aes(group = case_control_new),
     linewidth = 0.5,
     linetype = 1,
     type = "norm"
   ) +

  geom_point(size = 3.5) +
  scale_color_manual(values = case_cols) +
  theme_classic(base_size = 16) +
  theme(
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 20, face = "bold"),
    axis.title.y = element_text(size = 20, face = "bold"),
    plot.title = element_text(size = 24, hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(size = 22, hjust = 0.5),
    legend.title = element_blank(),
    legend.text = element_text(size = 18),
    legend.margin = margin(0, 0, 0, 0),

    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),

    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank()
  ) +
  labs(
    x = xlab,
    y = ylab,
    title = bquote ("Aitchison Distance"),
    subtitle = bquote("Adjusted " * italic(p) == .(get_adj_p("Aitchison Distance")))
  )


p_aitch


# ----------------------------------------------------------
# Generalized UniFrac PCoA manually from GUniFrac distance
# ----------------------------------------------------------

ord_guni <- cmdscale(guni_dist, eig = TRUE, k = 2)

guni_var <- round(ord_guni$eig / sum(ord_guni$eig) * 100, 1)

guni_df <- data.frame(
  Subject.ID = rownames(ord_guni$points),
  Axis1 = ord_guni$points[, 1],
  Axis2 = ord_guni$points[, 2]
) %>%
  left_join(
    merged_meta_r %>%
      dplyr::select(Subject.ID, case_control_new),
    by = "Subject.ID"
  )

p_guni <- ggplot(
  guni_df,
  aes(
    x = Axis1,
    y = Axis2,
    color = case_control_new
  )
) +
  stat_ellipse(
    aes(group = case_control_new),
    linewidth = 0.5,
    show.legend = FALSE
  ) +
  geom_point(size = 3.5) +
  scale_color_manual(values = case_cols) +
  theme_classic(base_size = 16) +
  theme(
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 20, face = "bold"),
    axis.title.y = element_text(size = 20, face = "bold"),
    plot.title = element_text(size = 24, hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(size = 22, hjust = 0.5),
    legend.title = element_blank (),
    legend.text = element_text(size = 18),
    legend.margin = margin(0, 0, 0, 0),
    
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    
    legend.background = element_rect(fill = "transparent", colour = NA),
    legend.box.background = element_blank(),
    legend.key = element_blank()
  ) +
  labs(
    #title = "PCoA using Generalized UniFrac",
    x = paste0("PCoA1 [", guni_var[1], "%]"),
    y = paste0("PCoA2 [", guni_var[2], "%]"),
    title = bquote("Generalized UniFrac"),
    subtitle = bquote("Adjusted " * italic(p) == .(get_adj_p("Generalized UniFrac")))#,
    #color = "Case/control"
  )

# Print plots
p_uni
p_guni
p_jac
p_wuni
p_bray
p_aitch

```

## Save beta diversity figures
```{r}

ggsave(
  filename = "Bray-Curtis.pdf",
  plot = p_bray,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)
ggsave(
  filename = "Bray-Curtis.png",
  plot = p_bray,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  bg = "transparent"
)

ggsave(
  filename = "UniFrac.pdf",
  plot = p_uni,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)

ggsave(
  filename = "UniFrac.png",
  plot = p_uni,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  bg = "transparent"
)

ggsave(
  filename = "WUniFrac.pdf",
  plot = p_wuni,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)
ggsave(
  filename = "WUniFrac.png",
  plot = p_wuni,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  bg = "transparent"
)

ggsave(
  filename = "GUniFrac.pdf",
  plot = p_guni,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)

ggsave(
  filename = "GUniFrac.png",
  plot = p_guni,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  bg = "transparent"
)

ggsave(
  filename = "Jaccard.pdf",
  plot = p_jac,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)

ggsave(
  filename = "Jaccard.png",
  plot = p_jac,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  bg = "transparent"
)

ggsave(
  filename = "Aitchison.pdf",
  plot = p_aitch,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)

ggsave(
  filename = "Aitchison.png",
  plot = p_aitch,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  bg = "transparent"
)
```
