---
title: "endo_vaginal_v9_pain_sensitivity_final"
author: "Cameron A. Gorzney, Stephen Johnson, Raunak Vijayakar, Michelle Bland"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries

```{r}
###r version 4.3.2

# # Install specific Bioconductor version
# install.packages("BiocManager")
# 
# BiocManager::install(version = "3.17")
# 
# # Then install packages
# BiocManager::install(c("phyloseq", "microbiome"))
```


```{r}
# packageVersion("")
library(vegan) #remotes::install_version("vegan", version = "2.6-10")
library(ggrepel)
library(readxl) #remotes::install_version("readxl", version = "1.4.5")  
library(phyloseq) #see above
library(ggplot2) #remotes::install_version("ggplot2", version = "4.0.0")  
library(hrbrthemes) #remotes::install_version("hrbrthemes", version = "0.8.7")  
library(ggsci) #remotes::install_version("ggsci", version = "3.2.0")  
library(dplyr) #remotes::install_version("dplyr", version = "1.1.4")  
library(microViz) #see below #needed to install "ggiraph" and "ComplexHeatmap"
# install.packages(
#   "microViz",
#   repos = c(davidbarnett = "https://david-barnett.r-universe.dev", getOption("repos"))
# )
library(microbiome) #see above
library(MicrobiomeStat) #remotes::install_version("MicrobiomeStat", version = "1.0")  
library(GUniFrac) #remotes::install_version("GUniFrac", version = "1.8")
library(picante) #remotes::install_version("picante", version = "1.8.2")
library(ggtext) #remotes::install_version("ggtext", version = "2.6-10")
library(tidyr)
library(purrr)
```

## Process samples

```{r}

setwd()


phylo = readRDS("data/Phylo_public_original_labels.rds")
meta = samdat_tbl(phylo)

#look at metadata
summarize_phyloseq(phylo)

table(meta$Source) #I thought this was a binary variable? We'll use only vaginal and rectal
table(meta$Plate, meta$case_control) #maybe include plate as covariate in models?

```

Add sample library size.

```{r}
depth = data.frame(lib_size = sample_sums(phylo))
depth$SampleID = rownames(depth)
depth$SampleID = sub('.', '', depth$SampleID)

phylo = ps_join(x = phylo, y = depth, by = "SampleID", type = "inner", match_sample_names = NULL)

sample_data(phylo)$subject = as.factor(sample_data(phylo)$subject)
```


```{r}
unique(meta$Source)
unique(meta$Amp.Status)

phylo_r = phylo |> ps_filter(Source == "vagina") 
temp = sample_data(phylo_r)
unique(temp$Subject.ID)

colSums(table(temp$Subject.ID, temp$case_control) != 0) #check sample size

phylo_r = phylo_r |> ps_filter(Amp.Status == "Amp")
meta_r = samdat_tbl(phylo_r)
unique(meta_r$Subject.ID)

colSums(table(meta_r$Subject.ID, meta_r$case_control) != 0) #check sample size

#list the subjects who were lost after filtering for amplification
unique(temp$Subject.ID[!(temp$Subject.ID %in% meta_r$Subject.ID)])

```

Merge and filter.

```{r}
merged_r = merge_samples(phylo_r, "subject", fun = sum)
# We use the sum function here instead of mean to get correct lib_size.
# The other numeric metadata are completely wrecked, as they would be with any function.

# Keep subject IDs from the merged phyloseq object
sample_data(merged_r)$Subject.ID = sample_names(merged_r)

# Add corrected patient demographics metadata
newdata = read_xlsx(
"data/Newdata_endo_pain.xlsx",col_names = T, sheet = 1)

splits = strsplit(newdata$`Study ID`, split = "-")
splits = lapply(splits, head, 2)
newdata$Subject.ID = sapply(splits, paste, collapse = "")


# Keep all clinical variables from the corrected demographics sheet
newdata = newdata |> 
  dplyr::rename(
    case_control_new = `case vs control`,
    paratubal_cysts_new = `Paratubal Cysts`,
    Gravidity_new = Graviditity,
    Age_new = `Age at surgery`,
    parity_status_new= `Parity status`,
    parity_new = Parity,
    BMI_new = BMI,
    stage_new = stage,
    Severity_new = Severity,
    Endosalpingiosis_new = Endosalpingiosis,
    Adenomyosis_new = Adenomyosis,
    Fibroids_new = Fibroids,
    pain_1st_level_new = pain_1st_level,
    abnormal_bleeding_new = Abnormal_bleeding
  ) |>
  dplyr::select(
    Subject.ID,
    case_control_new,
    BMI_new,
    Age_new,
    Gravidity_new,
    stage_new,
    Severity_new,
    parity_status_new,
    Endosalpingiosis_new,
    Adenomyosis_new,
    Fibroids_new,
    paratubal_cysts_new,
    parity_new,
    pain_1st_level_new,
    abnormal_bleeding_new,
    Race,
    Ethnicity
  ) |>
  dplyr::distinct(Subject.ID, .keep_all = TRUE)

merged_r = ps_join(merged_r, newdata, by = "Subject.ID", type = "left")

meta_pain <- samdat_tbl(merged_r)

pain_subject <- meta_pain %>%
  dplyr::select(Subject.ID, pain_1st_level_new) %>%
  dplyr::distinct() %>%
  dplyr::mutate(
    pain_score = as.numeric(pain_1st_level_new),
    pain_score = dplyr::case_when(
      is.na(pain_score) ~ NA_real_,
      pain_score == 99 ~ NA_real_,
      pain_score < 0 ~ NA_real_,
      pain_score > 10 ~ NA_real_,
      TRUE ~ pain_score
    ),
    pain_severity = dplyr::case_when(
      pain_score == 0 ~ "No pain",
      pain_score >= 1 & pain_score <= 3 ~ "Mild pain",
      pain_score >= 4 & pain_score <= 6 ~ "Moderate pain",
      pain_score >= 7 & pain_score <= 10 ~ "Severe pain",
      TRUE ~ NA_character_
    ),
    pain_severity = factor(
      pain_severity,
      levels = c("No pain", "Mild pain", "Moderate pain", "Severe pain")
    ),
    pain_binary = dplyr::case_when(
      pain_score <= 3 ~ "No/Mild",
      pain_score >= 4 ~ "Moderate/Severe",
      TRUE ~ NA_character_
    ),
    pain_binary = factor(
      pain_binary,
      levels = c("No/Mild", "Moderate/Severe")
    )
  ) %>%
  dplyr::filter(!is.na(pain_score)) %>%
  dplyr::select(Subject.ID, pain_score, pain_severity, pain_binary)

table(pain_subject$pain_score)
table(pain_subject$pain_severity)
table(pain_subject$pain_binary)


merged_r = ps_join(merged_r, pain_subject, by = "Subject.ID", type = "inner")

# Number of unique participants with BOTH microbiome + pain data
n_distinct(sample_data(merged_r)$Subject.ID)

# Check that corrected demographics are still present after join
merged_meta_check = samdat_tbl(merged_r)

colnames(merged_meta_check)

table(merged_meta_check$case_control_new, useNA = "ifany")
summary(merged_meta_check$BMI_new)
summary(merged_meta_check$Age_new)
summary(merged_meta_check$parity_new)
table(merged_meta_check$Severity_new, useNA = "ifany")


merged_r <- merged_r %>%
  ps_filter(lib_size > 2000) %>%
  tax_filter(min_prevalence = 2)

merged_meta_r = samdat_tbl(merged_r)

merged_meta_r$case_control_new = as.factor(merged_meta_r$case_control_new)
merged_meta_r$case_control_new = relevel(merged_meta_r$case_control_new, ref = "control")
merged_meta_r$Fibroids_new = as.factor(merged_meta_r$Fibroids_new)
merged_meta_r$paratubal_cysts_new = as.factor(merged_meta_r$paratubal_cysts_new)
merged_meta_r$Endosalpingiosis_new = as.factor(merged_meta_r$Endosalpingiosis_new)
merged_meta_r$Adenomyosis_new = as.factor(merged_meta_r$Adenomyosis_new)


#check for library size confounding
summary(merged_meta_r$lib_size)
hist(merged_meta_r$lib_size)
boxplot(merged_meta_r$lib_size ~ merged_meta_r$case_control_new, notch = F)
wilcox.test(merged_meta_r$lib_size ~ merged_meta_r$case_control_new)
boxplot(merged_meta_r$lib_size ~ merged_meta_r$pain_severity, notch = F)
kruskal.test(merged_meta_r$lib_size ~ merged_meta_r$pain_severity)


#rarefy
set.seed(123)
rare_r = rarefy_even_depth(merged_r, replace = F) #rarefy to min depth or 2k, whichever greater

```

### Checking other diagnoses against Pain Severity
```{r}
categorical_pvals <- tibble::tibble(
  Variable = c(
    "Adenomyosis",
    "Endosalpingiosis",
    "Fibroids",
    "Endometriosis",
    "Paratubal Cysts",
    "Abnormal Bleeding",
    "Race",
    "Ethnicity"
    
  ),
  p_value = c(
    fisher.test(table(merged_meta_r$Adenomyosis_new,
                      merged_meta_r$pain_severity))$p.value,

    fisher.test(table(merged_meta_r$Endosalpingiosis_new,
                      merged_meta_r$pain_severity))$p.value,

    fisher.test(table(merged_meta_r$Fibroids_new,
                      merged_meta_r$pain_severity))$p.value,
    
    fisher.test(table(merged_meta_r$case_control_new,
                      merged_meta_r$pain_severity))$p.value,

    fisher.test(table(merged_meta_r$paratubal_cysts_new,
                      merged_meta_r$pain_severity))$p.value,

    fisher.test(table(merged_meta_r$abnormal_bleeding_new,
                      merged_meta_r$pain_severity))$p.value,

    fisher.test(table(merged_meta_r$Race,
                      merged_meta_r$pain_severity))$p.value,

    fisher.test(table(merged_meta_r$Ethnicity,
                      merged_meta_r$pain_severity))$p.value
  )
) %>%
  dplyr::mutate(
    FDR = p.adjust(p_value, method = "BH")
  
)

categorical_pvals

### Continous Summary
continuous_summary <- merged_meta_r %>%
  dplyr::filter(
    !is.na(pain_severity),
    pain_severity %in% c("No pain", "Mild pain", "Moderate pain", "Severe pain")
  ) %>%
  dplyr::select(
    Severity_new,
    Age_new,
    BMI_new,
    Gravidity_new,
    parity_new
  ) %>%
  tidyr::pivot_longer(
    cols = c(Age_new, BMI_new, Gravidity_new, parity_new),
    names_to = "Variable",
    values_to = "Value"
  ) %>%
  dplyr::group_by(Variable, Severity_new) %>%
  dplyr::summarise(
    mean = mean(Value, na.rm = TRUE),
    sd = sd(Value, na.rm = TRUE),
    mean_sd = paste0(round(mean, 2), " ± ", round(sd, 2)),
    n = sum(!is.na(Value)),
    .groups = "drop"
  )

continuous_summary

continuous_pvals_pain <- merged_meta_r %>%
  dplyr::filter(!is.na(pain_severity)) %>%
  dplyr::select(
    pain_severity,
    Age_new,
    BMI_new,
    Gravidity_new,
    parity_new
  ) %>%
  tidyr::pivot_longer(
    cols = c(Age_new, BMI_new, Gravidity_new, parity_new),
    names_to = "Variable",
    values_to = "Value"
  ) %>%
  dplyr::group_by(Variable) %>%
  dplyr::summarise(
    p_value = kruskal.test(Value ~ pain_severity)$p.value,
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    FDR = p.adjust(p_value, method = "BH")
  )

continuous_pvals_pain

```

### Alpha Diversity for pain severity

```{r}
set.seed(123)

metric_labels <- c(
  Observed = "Observed ASVs",
  Chao1 = "Chao1 Richness",
  Shannon = "Shannon Diversity Index",
  InvSimpson = "Inverse Simpson Index"
)

tax <- as.data.frame(tax_table(rare_r), stringsAsFactors = FALSE)

plot_richness(rare_r, x="pain_severity", measures=c("Observed", "Chao1", "Shannon", "InvSimpson")) + geom_boxplot()

rich_r <- estimate_richness(
  rare_r,
  measures = c("Observed", "Chao1", "Shannon", "InvSimpson")
)

rich_r$pain_severity <- sample_data(rare_r)$pain_severity

rich_r <- rich_r %>%
  mutate(
    Subject.ID = sample_data(rare_r)$Subject.ID,
    pain_score = sample_data(rare_r)$pain_score,
    pain_severity = factor(
      sample_data(rare_r)$pain_severity,
      levels = c("No pain", "Mild pain", "Moderate pain", "Severe pain"),
      labels = c("No pain\n(n = 4)", "Mild pain\n(n = 7)", "Moderate pain\n(n = 5)", "Severe pain\n(n = 9)")
    ),
    case_control = sample_data(rare_r)$case_control_new
  )

table(rich_r$pain_severity)

alpha_results <- tibble(
  Metric = c("Observed", "Chao1", "Shannon", "InvSimpson"),
  Pvalue = c(
    kruskal.test(Observed ~ pain_severity, data = rich_r)$p.value,
    kruskal.test(Chao1 ~ pain_severity, data = rich_r)$p.value,
    kruskal.test(Shannon ~ pain_severity, data = rich_r)$p.value,
    kruskal.test(InvSimpson ~ pain_severity, data = rich_r)$p.value
  )
) %>%
  mutate(
    FDR = p.adjust(Pvalue, method = "fdr")#,
    #p_label = paste0("Kruskal-Wallis p = ", signif(Pvalue, 3))
  )

alpha_results

plot_alpha_metric <- function(metric_name) {
  
  p_value <- alpha_results %>%
    filter(Metric == metric_name) %>%
    pull(Pvalue)
  
  p_lab <- bquote("Kruskal-Wallis " * italic(p) == .(signif(p_value, 3)))
  
  ggplot(
    rich_r,
    aes(
      x = pain_severity,
      y = .data[[metric_name]],
      fill = pain_severity
    )
  ) +
    geom_boxplot(
      outlier.shape = NA,
      alpha = 0.75,
      color = "black"
    ) +
    geom_jitter(
      width = 0.12,
      size = 3,
      alpha = 0.85,
      color = "black"
    ) +
    scale_fill_manual(
      values = c(
        "No pain\n(n = 4)" = "grey",
        "Mild pain\n(n = 7)" = "#FDBB84",
        "Moderate pain\n(n = 5)" = "#EF6548",
        "Severe pain\n(n = 9)" = "#990000"
      )
    ) +
    theme_classic(base_size = 14) +
    theme(
      legend.position = "none",
      axis.text.x = element_text(angle = 45, hjust = 1, size = 18, face = "bold"),
      axis.text.y = element_text(size = 18),
      axis.title.x = element_text(size = 18),
      axis.title.y = element_text(size = 18, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 24, face = "bold"),
      panel.background = element_rect(fill = "transparent", colour = NA),
      plot.background = element_rect(fill = "transparent", colour = NA)
    ) +
    labs(
      x = "Pain Severity",
      y = metric_labels[[metric_name]],
      #title = paste0(metric_labels[[metric_name]], " by pain severity"),
      subtitle = p_lab
    )
}
p_observed <- plot_alpha_metric("Observed")
p_chao1 <- plot_alpha_metric("Chao1")
p_shannon <- plot_alpha_metric("Shannon")
p_invsimpson <- plot_alpha_metric("InvSimpson")

p_observed
p_chao1
p_shannon
p_invsimpson
```


### Save figures
```{r}
# Save figures

ggsave(
  filename = "Observed_alpha_diversity.png",
  plot = p_observed,
  path = fig_dir,
  width = 6,
  height = 5.5,
  units = "in",
  dpi = 600,
  bg = "transparent"
)

ggsave(
  filename = "Observed_alpha_diversity.pdf",
  plot = p_observed,
  path = fig_dir,
  width = 6,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)

ggsave(
  filename = "Chao1_alpha_diversity.png",
  plot = p_chao1,
  path = fig_dir,
  width = 6,
  height = 5.5,
  units = "in",
  dpi = 600,
  bg = "transparent"
)

ggsave(
  filename = "Chao1_alpha_diversity.pdf",
  plot = p_chao1,
  path = fig_dir,
  width = 6,
  height = 5.5,
  units = "in",
  dpi = 600,
  bg = "transparent"
)

ggsave(
  filename = "Shannon_alpha_diversity.png",
  plot = p_shannon,
  path = fig_dir,
  width = 6,
  height = 5.5,
  units = "in",
  dpi = 600,
  bg = "transparent"
)

ggsave(
  filename = "Shannon_alpha_diversity.pdf",
  plot = p_shannon,
  path = fig_dir,
  width = 6,
  height = 5.5,
  units = "in",
  dpi = 600,
  bg = "transparent"
)
ggsave(
  filename = "InvSimpson_alpha_diversity.png",
  plot = p_invsimpson,
  path = fig_dir,
  width = 6,
  height = 5.5,
  units = "in",
  dpi = 600,
  bg = "transparent"
)

ggsave(
  filename = "InvSimpson_alpha_diversity.pdf",
  plot = p_invsimpson,
  path = fig_dir,
  width = 6,
  height = 5.5,
  units = "in",
  dpi = 600,
  bg = "transparent"
)
```


Phylogenetic diversity from `picante`.

```{r}
set.seed(123)

pd_result <- pd(
  otu_table(rare_r),
  phy_tree(rare_r),
  include.root = TRUE
)

pd_df <- pd_result %>%
  as.data.frame()

pd_df$Subject.ID <- rownames(pd_df)

pd_df$pain_severity <- sample_data(rare_r)$pain_severity

pd_df <- pd_df %>%
  mutate(
    pain_severity = factor(
      pain_severity,
      levels = c("No pain", "Mild pain", "Moderate pain", "Severe pain"),
      labels = c("No pain\n(n = 4)", "Mild pain\n(n = 7)", "Moderate pain\n(n = 5)", "Severe pain\n(n = 9)")
    )
  )

pd_test <- kruskal.test(
  PD ~ pain_severity,
  data = pd_df
)

pd_test

pd_label <- bquote("Kruskal-Wallis " * italic(p) == .(signif(pd_test$p.value, 3)))

pd_plot <- ggplot(
  pd_df,
  aes(
    x = pain_severity,
    y = PD,
    fill = pain_severity
  )
) +
  geom_boxplot(
    outlier.shape = NA,
    alpha = 0.75,
    color = "black"
  ) +
  geom_jitter(
    width = 0.12,
    size = 3,
    alpha = 0.85,
    color = "black"
  ) +
  scale_fill_manual(
    values = c(
      "No pain\n(n = 4)" = "grey",
      "Mild pain\n(n = 7)" = "#FDBB84",
      "Moderate pain\n(n = 5)" = "#EF6548",
      "Severe pain\n(n = 9)" = "#990000"
    )
  ) +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 18, face = "bold"),
    axis.text.y = element_text(size = 18),
      axis.title.x = element_text(size = 18),
      axis.title.y = element_text(size = 18, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 24, face = "bold"),
      panel.background = element_rect(fill = "transparent", colour = NA),
      plot.background = element_rect(fill = "transparent", colour = NA)
    ) +
  labs(
    x = "Pain Severity",
    y = "Faith's Phylogenetic Diversity",
    #title = "Phylogenetic diversity by pain severity",
    subtitle = pd_label
  )
pd_plot
```

# Save Figures
```{r}
ggsave(
  filename = "Faith_PD.png",
  plot = pd_plot,
  path = fig_dir,
  width = 6,
  height = 5.5,
  units = "in",
  dpi = 600,
  bg = "transparent"
)

ggsave(
  filename = "Faith_PD.pdf",
  plot = pd_plot,
  path = fig_dir,
  width = 6,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)
```

### New beta diversity for pain severity, with plots

```{r}
set.seed(123)

# Match metadata to rare_r sample order
merged_meta_r <- merged_meta_r %>%
  filter(Subject.ID %in% sample_names(rare_r)) %>%
  arrange(match(Subject.ID, sample_names(rare_r)))

merged_meta_r$pain_severity <- sample_data(rare_r)[merged_meta_r$Subject.ID, ]$pain_severity

merged_meta_r <- merged_meta_r %>%
  mutate(
    pain_severity = factor(
      pain_severity,
      levels = c("No pain", "Mild pain", "Moderate pain", "Severe pain"),
      labels = c(
        "No pain (n = 4)",
        "Mild pain (n = 7)",
        "Moderate pain (n = 5)",
        "Severe pain (n = 9)"
      )
    )
  )

table(merged_meta_r$pain_severity, useNA = "ifany")

stopifnot(all(merged_meta_r$Subject.ID == sample_names(rare_r)))

# ----------------------------------------------------------
# Build distance matrices
# ----------------------------------------------------------

tree_r <- phy_tree(rare_r)

otu_r <- as(otu_table(rare_r), "matrix")

if (taxa_are_rows(rare_r)) {
  otu_r <- t(otu_r)
}

otu_r <- otu_r[merged_meta_r$Subject.ID, ]

unifracs <- GUniFrac(
  otu.tab = otu_r,
  tree = tree_r,
  alpha = c(0, 0.5, 1)
)$unifracs

uni_dist <- as.dist(unifracs[, , "d_UW"])
guni_dist <- as.dist(unifracs[, , "d_0.5"])
wuni_dist <- as.dist(unifracs[, , "d_1"])
jac_dist <- vegdist(otu_r, method = "jaccard", binary = TRUE)
bray_dist <- vegdist(otu_r, method = "bray")

### Prepare pseudocounts for aitchison
# Genus-level non-rarefied counts
merged_r_genus <- tax_glom(merged_r, taxrank = "Genus")

genus_counts <- as.data.frame(otu_table(merged_r_genus))

# Make sure rows = subjects and columns = taxa
if (taxa_are_rows(merged_r_genus)) {
  genus_counts <- as.data.frame(t(genus_counts))
}

# Add clean genus names as column names
genera <- as.data.frame(tax_table(merged_r_genus))
genus_names <- as.character(genera$Genus)
genus_names[is.na(genus_names) | genus_names == ""] <- "unclassified"
genus_names <- make.unique(genus_names)

colnames(genus_counts) <- genus_names

# CLR transform
genus_counts_pseudo <- genus_counts + 1
clr_mat <- compositions::clr(genus_counts_pseudo)

# Aitchison PCA
pca1 <- prcomp(clr_mat, center = TRUE, scale. = FALSE)
summary(pca1)

scores <- as.data.frame(pca1$x)
scores$Subject.ID <- rownames(scores)

# Aitchison distance
aitch <- vegan::vegdist(clr_mat, method = "euclidean")
# ----------------------------------------------------------
# PERMANOVA
# ----------------------------------------------------------

uni_test <- adonis3(uni_dist ~ pain_severity, data = merged_meta_r)
guni_test <- adonis3(guni_dist ~ pain_severity, data = merged_meta_r)
jac_test <- adonis3(jac_dist ~ pain_severity, data = merged_meta_r)
wuni_test <- adonis3(wuni_dist ~ pain_severity, data = merged_meta_r)
bray_test <- adonis3(bray_dist ~ pain_severity, data = merged_meta_r)
aitch_test <- adonis3(aitch ~ pain_severity, data = merged_meta_r)

beta_results <- data.frame(
  Metric = c(
    "UniFrac",
    "Generalized UniFrac",
    "Jaccard",
    "Weighted UniFrac",
    "Bray-Curtis",
    "Aitchison Distance"
  ),
  R2 = c(
    uni_test$aov.tab$R2[1],
    guni_test$aov.tab$R2[1],
    jac_test$aov.tab$R2[1],
    wuni_test$aov.tab$R2[1],
    bray_test$aov.tab$R2[1],
    aitch_test$aov.tab$R2[1]
  ),
  Pvalue = c(
    uni_test$aov.tab$`Pr(>F)`[1],
    guni_test$aov.tab$`Pr(>F)`[1],
    jac_test$aov.tab$`Pr(>F)`[1],
    wuni_test$aov.tab$`Pr(>F)`[1],
    bray_test$aov.tab$`Pr(>F)`[1],
    aitch_test$aov.tab$`Pr(>F)`[1]
  )
)

beta_results$FDR <- p.adjust(beta_results$Pvalue, method = "fdr")

beta_results

# ----------------------------------------------------------
# Plot settings
# ----------------------------------------------------------

pain_cols <- c(
  "No pain (n = 4)" = "grey",
  "Mild pain (n = 7)" = "#FDBB84",
  "Moderate pain (n = 5)" = "#EF6548",
  "Severe pain (n = 9)" = "#990000"
)

get_p <- function(metric_name) {
  signif(beta_results$Pvalue[beta_results$Metric == metric_name], 3)
}

get_adj_p <- function(metric_name) {
  signif(beta_results$FDR[beta_results$Metric == metric_name], 3)
}

# ----------------------------------------------------------
# PCoA plots from phyloseq-supported distances
# ----------------------------------------------------------
sample_data(rare_r)$pain_severity <- factor(
  sample_data(rare_r)$pain_severity,
  levels = c("No pain", "Mild pain", "Moderate pain", "Severe pain"),
  labels = c(
    "No pain (n = 4)",
    "Mild pain (n = 7)",
    "Moderate pain (n = 5)",
    "Severe pain (n = 9)"
  )
)

ord_uni <- ordinate(rare_r, "PCoA", "unifrac")

p_uni <- plot_ordination(rare_r, ord_uni, color = "pain_severity") +
  stat_ellipse(
    aes(group = pain_severity),
    linewidth = 0.5,
    show.legend = FALSE
  ) +
  geom_point(size = 3.5) +
  scale_color_manual(values = pain_cols) +
  theme_classic(base_size = 16) +
 theme(
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 24),
    plot.subtitle = element_text(size = 22, hjust = 0.5),
    legend.title = element_blank(),
    legend.text = element_text(size = 18),
    legend.margin = margin(0, 0, 0, 0),
    
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank()
  ) +
  labs(
    #title = "PCoA using UniFrac",
    x = paste0("PCoA1 [", round(ord_uni$values$Relative_eig[1] * 100, 1), "%]"),
    y = paste0("PCoA2 [", round(ord_uni$values$Relative_eig[2] * 100, 1), "%]"),
    title = bquote("UniFrac"),
    subtitle = bquote("Adjusted " * italic(p) == .(get_adj_p("UniFrac"))),
    color = "Pain Severity"
  )

ord_jac <- ordinate(rare_r, "PCoA", "jaccard")

p_jac <- plot_ordination(rare_r, ord_jac, color = "pain_severity") +
  stat_ellipse(
    aes(group = pain_severity),
    linewidth = 0.5,
    show.legend = FALSE
  ) +
  geom_point(size = 3.5) +
  scale_color_manual(values = pain_cols) +
  theme_classic(base_size = 16) +
  theme(
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 24),
    plot.subtitle = element_text(size = 22, hjust = 0.5),
    legend.title = element_blank(),
    legend.text = element_text(size = 18),
    legend.margin = margin(0, 0, 0, 0),
    
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank()
  ) +
  labs(
    #title = "PCoA using Jaccard",
    x = paste0("PCoA1 [", round(ord_jac$values$Relative_eig[1] * 100, 1), "%]"),
    y = paste0("PCoA2 [", round(ord_jac$values$Relative_eig[2] * 100, 1), "%]"),
    title = bquote("Jaccard"),
    subtitle = bquote("Adjusted " * italic(p) == .(get_adj_p("Jaccard"))),
    color = "Pain Severity"
  )

ord_wuni <- ordinate(rare_r, "PCoA", "wunifrac")

p_wuni <- plot_ordination(rare_r, ord_wuni, color = "pain_severity") +
  stat_ellipse(
    aes(group = pain_severity),
    linewidth = 0.5,
    show.legend = FALSE
  ) +
  geom_point(size = 3.5) +
  scale_color_manual(values = pain_cols) +
  theme_classic(base_size = 16) +
  theme(
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 24),
    plot.subtitle = element_text(size = 22, hjust = 0.5),
    legend.title = element_blank(),
    legend.text = element_text(size = 18),
    legend.margin = margin(0, 0, 0, 0),
    
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank()
  ) +
  labs(
    #title = "PCoA using Weighted UniFrac",
    x = paste0("PCoA1 [", round(ord_wuni$values$Relative_eig[1] * 100, 1), "%]"),
    y = paste0("PCoA2 [", round(ord_wuni$values$Relative_eig[2] * 100, 1), "%]"),
    title = bquote("Weighted UniFrac"),
    subtitle = bquote("Adjusted " * italic(p) == .(get_adj_p("Weighted UniFrac"))),
    color = "Pain Severity"
  )

ord_bray <- ordinate(rare_r, "PCoA", "bray")

p_bray <- plot_ordination(rare_r, ord_bray, color = "pain_severity") +
  stat_ellipse(
    aes(group = pain_severity),
    linewidth = 0.5,
    show.legend = FALSE
  ) +
  geom_point(size = 3.5) +
  scale_color_manual(values = pain_cols) +
  theme_classic(base_size = 16) +
  theme(
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 24),
    plot.subtitle = element_text(size = 22, hjust = 0.5),
    legend.title = element_blank(),
    legend.text = element_text(size = 18),
    legend.margin = margin(0, 0, 0, 0),
    
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank()
  ) +
  labs(
    #title = "PCoA using Bray-Curtis",
    x = paste0("PCoA1 [", round(ord_bray$values$Relative_eig[1] * 100, 1), "%]"),
    y = paste0("PCoA2 [", round(ord_bray$values$Relative_eig[2] * 100, 1), "%]"),
    title = bquote ("Bray-Curtis"),
    subtitle = bquote("Adjusted " * italic(p) == .(get_adj_p("Bray-Curtis"))),
    color = "Pain Severity"
  )


# PCA scores
scores <- as.data.frame(pca1$x)
scores$Subject.ID <- rownames(scores)

# Verify IDs match metadata
setequal(scores$Subject.ID, merged_meta_r$Subject.ID)


# Add pain severity to metadata
scores <- scores %>%
  dplyr::left_join(
    merged_meta_r %>%
      dplyr::select(Subject.ID, pain_severity),
    by = "Subject.ID"
  )

table(scores$pain_severity, useNA = "always")


# Axis labels
var_exp <- (pca1$sdev^2) / sum(pca1$sdev^2)

xlab <- paste0("PC1 [", round(100 * var_exp[1], 1), "%]")
ylab <- paste0("PC2 [", round(100 * var_exp[2], 1), "%]")

#Add Aitchison axes and clusters directly to merged_meta_r
merged_meta_r <- merged_meta_r %>%
dplyr::select(
     -any_of(c(
       "PC1",
       "PC2"
     ))
   ) %>%
   dplyr::left_join(
     scores %>%
       dplyr::select(
         Subject.ID,
         PC1,
         PC2,
       ),
     by = "Subject.ID"
   )

# Aitchison PCA plot
p_aitch <- ggplot(
   merged_meta_r,
   aes(
     x = PC1,
     y = PC2,
     color = pain_severity
   )
 ) +
 
   stat_ellipse(
    aes(group = pain_severity),
    linewidth = 0.5,
    show.legend = FALSE
  ) +

  geom_point(size = 3.5) +
  scale_color_manual(values = pain_cols) +
  theme_classic(base_size = 16) +
  theme(
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 24),
    plot.subtitle = element_text(size = 22, hjust = 0.5),
    legend.title = element_blank(),
    legend.text = element_text(size = 18),
    legend.margin = margin(0, 0, 0, 0),

    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),

    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank()
  ) +
  labs(
    x = xlab,
    y = ylab,
    title = bquote("Aitchison Distance"),
    subtitle = bquote("Adjusted " * italic(p) == .(get_adj_p("Aitchison Distance")))
  )


p_aitch

# ----------------------------------------------------------
# Generalized UniFrac PCoA manually from GUniFrac distance
# ----------------------------------------------------------

ord_guni <- cmdscale(guni_dist, eig = TRUE, k = 2)

guni_var = round(ord_guni$eig/sum(ord_guni$eig)*100, 1)

guni_df <- data.frame(
  Subject.ID = rownames(ord_guni$points),
  Axis1 = ord_guni$points[, 1],
  Axis2 = ord_guni$points[, 2]
) %>%
  left_join(
    merged_meta_r %>%
      dplyr::select(Subject.ID, pain_severity),
    by = "Subject.ID"
  )

p_guni <- ggplot(
  guni_df,
  aes(
    x = Axis1,
    y = Axis2,
    color = pain_severity
  )
) +
  stat_ellipse(
    aes(group = pain_severity),
    linewidth = 0.5,
    show.legend = FALSE
  ) +
  geom_point(size = 3.5) +
  scale_color_manual(values = pain_cols) +
  theme_classic(base_size = 16) +
  theme(
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    #plot.title = element_text(size = 22, hjust = 0.5),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 24),
    plot.subtitle = element_text(size = 22, hjust = 0.5),
    legend.title = element_blank(),
    legend.text = element_text(size = 18),
    legend.margin = margin(0, 0, 0, 0),
    
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank()
  ) +
  labs(
    #title = "PCoA using Generalized UniFrac",
    x = paste0("PCoA1 [", guni_var[1], "%]"),
    y = paste0("PCoA2 [", guni_var[2], "%]"),
    title = bquote("Generalized UniFrac"),
    subtitle = bquote("Adjusted " * italic(p) == .(get_adj_p("Generalized UniFrac"))),
    color = "Pain Severity"
  )

# Print individual plots (raw p-values)
p_uni
p_guni #not sure why guni didn't have %var on axes. -fixed it above.
p_jac
p_wuni
p_bray
p_aitch
```
### Save beta-diversity plots
```{r}
ggsave(
  filename = "Bray-Curtis.pdf",
  plot = p_bray,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)
ggsave(
  filename = "Bray-Curtis.png",
  plot = p_bray,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  bg = "transparent"
)

ggsave(
  filename = "UniFrac.pdf",
  plot = p_uni,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)

ggsave(
  filename = "UniFrac.png",
  plot = p_uni,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  bg = "transparent"
)

ggsave(
  filename = "WUniFrac.pdf",
  plot = p_wuni,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)
ggsave(
  filename = "WUniFrac.png",
  plot = p_wuni,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  bg = "transparent"
)

ggsave(
  filename = "GUniFrac.pdf",
  plot = p_guni,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)

ggsave(
  filename = "GUniFrac.png",
  plot = p_guni,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  bg = "transparent"
)

ggsave(
  filename = "Jaccard.pdf",
  plot = p_jac,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)

ggsave(
  filename = "Jaccard.png",
  plot = p_jac,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  bg = "transparent"
)

ggsave(
  filename = "Aitch.pdf",
  plot = p_aitch,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)

```

#### PAM clustering

##### Bray-Curtis for pain severity

Calculate BC dissimilarity on rarefied data, then cluster with PAM, k selected using mean silhouette width.
```{r}
#determine number of clusters
library(cluster) #remotes::install_version("cluster", version = "2.1.8.1")
library(scales) #remotes::install_version("scales", version = "1.4.0")

#use sil to choose number of clusters
set.seed(123)
k.max <- 10

SIL = sapply(2:k.max, function(i) {
  results.coef.pam =  pam(bray_dist, i, nstart=50, diss = T)
  mean(silhouette(results.coef.pam$clustering, dist = bray_dist)[,"sil_width"]) })

plot(c(0,SIL), type="b", main = "Selecting k PAM clusters", ylab = "Mean Silhouette Width", xlab = "Clusters")


```

Visualize clusters and pain severity status on PCoA. 

```{r}
set.seed(123)
pam_bc = pam(bray_dist, 4, nstart = 50, diss = T)

# ord = cmdscale(bray_dist, 4, eig = TRUE)
# plot(ord$points, col = clusters, pch = 19, xlab="PCoA1", ylab="PCoA2")

sample_data(rare_r)$Cluster = as.factor(pam_bc$clustering) #add clustering information to ps object's sample data

ord = ordinate(rare_r, "PCoA", "bray")
plot_ordination(rare_r, ord, color = "pain_severity", shape = "Cluster") +
  geom_point(size = 3) + 
  labs(title = "PCoA using Bray-Curtis") +
  theme_ipsum() +
  scale_color_bmj()
```

Correlate rarefied abundances with PCoA axes to look for gradients.
```{r}
#save axes
bc_axis.1 = ord$vectors[,1]
bc_axis.2 = ord$vectors[,2]
```

### Merge BC axis with merged_meta_r 
```{r}
bc_axis1.df <- data.frame(
  Subject.ID = names(bc_axis.1),
  bray_pcoa1 = as.numeric(bc_axis.1)
)

merged_meta_r <- merged_meta_r %>%
  dplyr::select(-any_of("bray_pcoa1")) %>%
  left_join(bc_axis1.df, by = "Subject.ID")

sum(is.na(merged_meta_r$bray_pcoa1))
head(merged_meta_r[, c("Subject.ID", "bray_pcoa1")])
all(bc_axis1.df$Subject.ID %in% merged_meta_r$Subject.ID)
```
### CLR, Aitchison PCA, PAM clusters

```{r}
library(zCompositions)
library(compositions)
library(cluster)


# Select k for PAM clustering
set.seed(123)
k.max <- 10

SIL <- sapply(2:k.max, function(i) {
  pam_fit <- pam(aitch, i, nstart = 50, diss = TRUE)
  mean(cluster::silhouette(pam_fit$clustering, dist = aitch)[, "sil_width"])
})

aitch_plot <- plot(
  2:k.max,
  SIL,
  type = "b",
  main = "Selecting k PAM clusters",
  ylab = "Mean Silhouette Width",
  xlab = "Clusters"
)


sil_df <- data.frame(
  Clusters = 2:k.max,
  Mean_Silhouette = SIL
)

p_sil <- ggplot(
  sil_df,
  aes(x = Clusters, y = Mean_Silhouette)
) +
  geom_line() +
  geom_point(size = 3) +
  theme_classic(base_size = 16) +
  theme(
    axis.text = element_text(size = 18, color = "black"),
    axis.title = element_text(size = 18, face = "bold"),
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA)
  ) +
  labs(
    x = "Clusters",
    y = "Mean Silhouette Width"
  )

p_sil
```

###Save
```{r}
ggsave(
  filename = "Aitch_SIL_plot.pdf",
  plot = p_sil,
  path = fig_dir,
  width = 6,
  height = 5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)
```

# PAM clustering using k = 2 --> identified 2vclusters after updating samples
```{r}
set.seed(123)
pam_aitch <- pam(aitch, 2, nstart = 50, diss = TRUE)

cluster_df <- data.frame(
  Subject.ID = rownames(clr_mat),
  Cluster_Aitchison = factor(pam_aitch$clustering)
)

scores <- scores %>%
  dplyr::left_join(cluster_df, by = "Subject.ID")

# Axis labels
var_exp <- (pca1$sdev^2) / sum(pca1$sdev^2)

xlab <- paste0("PC1 [", round(100 * var_exp[1], 1), "%]")
ylab <- paste0("PC2 [", round(100 * var_exp[2], 1), "%]")

# Add Aitchison axes and clusters directly to merged_meta_r
merged_meta_r <- merged_meta_r %>%
  dplyr::select(
    -any_of(c(
      "PC1",
      "PC2",
      "Cluster_Aitchison"
    ))
  ) %>%
  dplyr::left_join(
    scores %>%
      dplyr::select(
        Subject.ID,
        PC1,
        PC2,
        Cluster_Aitchison
      ),
    by = "Subject.ID"
  )

# Keep CLR taxa table available for later taxa/heatmap analyses
clr_df <- as.data.frame(clr_mat)
clr_df$Subject.ID <- rownames(clr_df)

pain_cols_n <- c(
  "No pain (n = 4)" = "grey",
  "Mild pain (n = 7)" = "#FDBB84",
  "Moderate pain (n = 5)" = "#EF6548",
  "Severe pain (n = 9)" = "#990000"
)

merged_meta_r <- merged_meta_r %>%
  mutate(
    pain_severity_plot = factor(
      as.character(pain_severity),
      levels = c(
        "No pain (n = 4)",
        "Mild pain (n = 7)",
        "Moderate pain (n = 5)",
        "Severe pain (n = 9)"
      ),
      ordered = TRUE
    ),
    pain_severity_clean = factor(
      dplyr::case_when(
        as.character(pain_severity) == "No pain (n = 4)" ~ "No pain",
        as.character(pain_severity) == "Mild pain (n = 7)" ~ "Mild pain",
        as.character(pain_severity) == "Moderate pain (n = 5)" ~ "Moderate pain",
        as.character(pain_severity) == "Severe pain (n = 9)" ~ "Severe pain"
      ),
      levels = c("No pain", "Mild pain", "Moderate pain", "Severe pain"),
      ordered = TRUE
    )
  )

# merged_meta_r$Cluster_Aitchison <- factor(
#   merged_meta_r$Cluster_Aitchison,
#   levels = c(1, 2, 3),
#   labels = c("Cluster 1", "Cluster 2", "Cluster 3")
# )

# Plot Aitchison PCA by pain severity and cluster
aitch_plot <- ggplot(
  merged_meta_r,
  aes(
    x = PC1,
    y = PC2,
    color = pain_severity,
    shape = Cluster_Aitchison
  )
) +
  
  stat_ellipse(
    data = merged_meta_r,
    aes(group = Cluster_Aitchison),
    linewidth = 0.5,
    linetype = 2,
    #color = "black",
    type = "norm"
  ) +
  
  geom_point(size = 4) +
  scale_color_manual(values = pain_cols_n) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    plot.title = element_text(size = 24, hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(size = 22, hjust = 0.5),
    legend.title = element_text(size = 18),
    legend.text = element_text(size = 18),
    legend.margin = margin(0, 0, 0, 0),
    
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank()
  ) +
  labs(
    title = "Aitchison Distance",
    x = xlab,
    y = ylab,
    color = "Pain Severity",
    shape = "Aitchison Cluster"
  )

aitch_plot
sum(is.na(merged_meta_r$PC1))
sum(is.na(merged_meta_r$Cluster_Aitchison))
all(scores$Subject.ID %in% merged_meta_r$Subject.ID)

```

### Save Clusters
```{r}
ggsave(
  filename = "Aitchison_pain.png",
  plot = aitch_plot,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  bg = "transparent"
)

ggsave(
  filename = "Aitchison_pain.pdf",
  plot = aitch_plot,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)

```

### Fisher's test for cluster based on pain severity
```{r}
fisher.test(
  table(
    merged_meta_r$Cluster_Aitchison,
    merged_meta_r$pain_severity
  )
)

tab <- table(
  merged_meta_r$Cluster_Aitchison,
  merged_meta_r$pain_severity
)

fisher.test(tab)

pairwise_results <- list()

pain_levels <- levels(merged_meta_r$pain_severity)

for(i in 1:(length(pain_levels)-1)) {
  for(j in (i+1):length(pain_levels)) {
    
    temp <- merged_meta_r %>%
      filter(
        pain_severity %in% c(
          pain_levels[i],
          pain_levels[j]
        )
      )
    
    tab <- table(
      temp$Cluster_Aitchison,
      temp$pain_severity
    )
    
    res <- fisher.test(tab)
    
    pairwise_results[[paste(
      pain_levels[i],
      "vs",
      pain_levels[j]
    )]] <- res$p.value
  }
}

pairwise_results

p.adjust(
  unlist(pairwise_results),
  method = "fdr"
)

```

## Graph cluster counts and Fisher's results
```{r}
cluster_counts <- merged_meta_r %>%
  mutate(
    pain_severity_plot = case_when(
      grepl("^No pain", as.character(pain_severity)) ~ "No pain (n = 4)",
      grepl("^Mild pain", as.character(pain_severity)) ~ "Mild pain (n = 7)",
      grepl("^Moderate pain", as.character(pain_severity)) ~ "Moderate pain (n = 5)",
      grepl("^Severe pain", as.character(pain_severity)) ~ "Severe pain (n = 9)",
      TRUE ~ NA_character_
    ),
    pain_severity_plot = factor(
      pain_severity_plot,
      levels = c(
        "No pain (n = 4)",
        "Mild pain (n = 7)",
        "Moderate pain (n = 5)",
        "Severe pain (n = 9)"
      )
    ),
    Cluster_Aitchison = factor(
      Cluster_Aitchison,
      levels = c(1, 2),
      labels = c("Cluster 1", "Cluster 2")
    )
  ) %>%
  filter(!is.na(pain_severity_plot)) %>%
  count(Cluster_Aitchison, pain_severity_plot, .drop = FALSE) %>%
  group_by(Cluster_Aitchison) %>%
  mutate(
    prop = n / sum(n),
    label = ifelse(n == 0, "", as.character(n))
  ) %>%
  ungroup()

p.adjust(
  unlist(pairwise_results),
  method = "fdr"
)


cluster_counts <- merged_meta_r %>%
  mutate(
    pain_severity_plot = case_when(
      grepl("^No pain", as.character(pain_severity)) ~ "No pain",
      grepl("^Mild pain", as.character(pain_severity)) ~ "Mild",
      grepl("^Moderate pain", as.character(pain_severity)) ~ "Moderate",
      grepl("^Severe pain", as.character(pain_severity)) ~ "Severe",
      TRUE ~ NA_character_
    ),
    pain_severity_plot = factor(
      pain_severity_plot,
      levels = c("No pain", "Mild", "Moderate", "Severe")
    ),
    Cluster_Aitchison = factor(
      Cluster_Aitchison,
      levels = c(1, 2),
      labels = c("Cluster 1", "Cluster 2")
    )
  ) %>%
  filter(!is.na(pain_severity_plot), !is.na(Cluster_Aitchison)) %>%
  count(Cluster_Aitchison, pain_severity_plot, .drop = FALSE) %>%
  mutate(label = ifelse(n == 0, "", as.character(n)))

clusters_pain_counts <- ggplot(
  cluster_counts,
  aes(
    x = Cluster_Aitchison,
    y = n,
    fill = pain_severity_plot
  )
) +
  geom_col(width = 0.85, color = "white") +
  geom_text(
    aes(label = label),
    position = position_stack(vjust = 0.5),
    size = 6,
    color = "black"
  ) +
  scale_fill_manual(
    values = c(
      "No pain" = "#737373",
      "Mild" = "#FDBB84",
      "Moderate" = "#EF6548",
      "Severe" = "#990000"
    ),
    drop = FALSE
  ) +
  scale_y_continuous(
    breaks = seq(0, 21, by = 3),
    expand = expansion(mult = c(0, 0.05))
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 18, face = "bold"),
    axis.title.y = element_text(size = 18, face = "bold"),
    legend.title = element_text(size = 18, face = "bold"),
    legend.text = element_text(size = 16),
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    
    legend.background = element_blank(),
    legend.box.background = element_blank(),
    legend.key = element_blank()
    
  ) +
  labs(
    x = NULL,
    y = "Number of Participants",
    fill = "Pain Severity"
  )

clusters_pain_counts

## Add fisher's p-value to the plot
fisher_p <- fisher.test(
  table(
    merged_meta_r$Cluster_Aitchison,
    merged_meta_r$pain_severity
  )
)$p.value

clusters_pain_counts <- clusters_pain_counts +
  labs(
    subtitle = bquote("Fisher's Exact " * italic(p) == .(signif(fisher_p, 3)))
  ) +
  theme(
    plot.subtitle = element_text(size = 20, hjust = 0.5, face = "bold")
  )

clusters_pain_counts

## Pairwise comparisons
pairwise_df <- tibble::tibble(
  comparison = names(pairwise_results),
  p_value = unlist(pairwise_results),
  fdr = p.adjust(unlist(pairwise_results), method = "fdr")
) %>%
  dplyr::mutate(
    p_label = paste0("p = ", signif(p_value, 3)),
    fdr_label = paste0("FDR = ", signif(fdr, 3))
  )

pairwise_df


```

# Save 
```{r}
ggsave(
  filename = "aitch_pain_clusters.png",
  plot = clusters_pain_counts,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  dpi = 600,
  bg = "transparent"
)

ggsave(
  filename = "aitch_pain_clusters.pdf",
  plot = clusters_pain_counts,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)
```
### Secondary cluster analayses...zicoseq 

```{r, eval =F}
############################################################
# Secondary analysis: characterize Aitchison clusters

############################################################

library(GUniFrac)
library(microeco)
library(phyloseq)
library(tidyverse)
library(ggrepel)
library(pheatmap)
library(ggpubr)

# Make sure cluster is a factor
merged_meta_r <- merged_meta_r %>%
  mutate(
    Cluster_Aitchison = factor(Cluster_Aitchison)
  )

############################################################
# Part 1. ZicoSeq DAA by Aitchison cluster
############################################################

# Genus-level count matrix
otu_cluster <- as.data.frame(otu_table(merged_r_genus))

# ZicoSeq expects taxa x samples
if (!taxa_are_rows(merged_r_genus)) {
  otu_cluster <- t(otu_cluster)
}

otu_cluster <- as.matrix(otu_cluster)
storage.mode(otu_cluster) <- "numeric"

# Match metadata to OTU columns
meta_cluster <- merged_meta_r %>%
  filter(
    Subject.ID %in% colnames(otu_cluster),
    !is.na(Cluster_Aitchison)
  ) %>%
  arrange(match(Subject.ID, colnames(otu_cluster))) %>%
  as.data.frame()

rownames(meta_cluster) <- meta_cluster$Subject.ID

otu_cluster <- otu_cluster[, meta_cluster$Subject.ID]

# Optional analysis-specific filter
otu_cluster <- otu_cluster[rowSums(otu_cluster > 0) >= 3, ]

# Checks
dim(otu_cluster)
nrow(meta_cluster)
all(colnames(otu_cluster) == rownames(meta_cluster))

set.seed(123)

zico_cluster <- ZicoSeq(
  meta.dat = meta_cluster,
  feature.dat = otu_cluster,
  grp.name = "Cluster_Aitchison",
  adj.name = NULL,
  feature.dat.type = "count",
  is.winsor = TRUE,
  outlier.pct = 0.03,
  winsor.end = "top",
  is.post.sample = FALSE,
  link.func = list(function(x) x^0.5),
  stats.combine.func = max,
  perm.no = 999,
  strata = NULL,
  ref.pct = 0.5,
  stage.no = 6,
  excl.pct = 0.2,
  is.fwer = FALSE,
  verbose = TRUE,
  return.feature.dat = TRUE
)

# Results table
zico_cluster_results <- data.frame(
  Taxon = names(zico_cluster$p.raw),
  p.raw = as.numeric(zico_cluster$p.raw),
  p.adj.fdr = as.numeric(zico_cluster$p.adj.fdr),
  R2 = as.numeric(zico_cluster$R2)
) %>%
  arrange(p.adj.fdr)

# Add taxonomy
tax_df_cluster <- as.data.frame(tax_table(merged_r_genus))
tax_df_cluster$Taxon <- rownames(tax_df_cluster)

zico_cluster_results <- zico_cluster_results %>%
  left_join(tax_df_cluster, by = "Taxon") %>%
  mutate(
    Genus_clean = ifelse(
      is.na(Genus) | Genus == "",
      Taxon,
      stringr::str_replace_all(Genus, "_", " ")
    ),
    Genus_clean = case_when(
      Taxon == "213ebca51a1d1b76ccd1456820b11a96548261a1" ~ "Eggerthellaceae unclassified",
      Taxon == "d7d0c7358e32837b6121cfd76b27f13f2a1edb75" ~ "Enterobacteriaceae unclassified",
      TRUE ~ Genus_clean
    )
  )
sig_zico_cluster <- zico_cluster_results %>%
  filter(p.adj.fdr < 0.05) 

sig_zico_cluster


############################################################
# Part 2. Cluster-defining taxa heatmap
############################################################

# CLR matrix with original taxon IDs
clr_cluster <- as.data.frame(clr_mat)
colnames(clr_cluster) <- taxa_names(merged_r_genus)
clr_cluster$Subject.ID <- rownames(clr_cluster)

clr_cluster_meta <- clr_cluster %>%
  left_join(
    merged_meta_r %>%
      dplyr::select(Subject.ID, Cluster_Aitchison),
    by = "Subject.ID"
  )

############################################################
# Part 3. Identify which Cluster each sig taxa is highest in
############################################################
if (nrow(sig_zico_cluster) > 0) {
  
  cluster_taxa_medians <- clr_cluster_meta %>%
    dplyr::select(
      Subject.ID,
      Cluster_Aitchison,
      all_of(sig_zico_cluster$Taxon)
    ) %>%
    pivot_longer(
      cols = all_of(sig_zico_cluster$Taxon),
      names_to = "Taxon",
      values_to = "CLR"
    ) %>%
    group_by(Taxon, Cluster_Aitchison) %>%
    summarise(
      Median_CLR = median(CLR, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    left_join(
      tax_df_cluster %>%
        dplyr::select(Taxon, Genus),
      by = "Taxon"
    ) %>%
    mutate(
      Genus_clean = ifelse(
        is.na(Genus) | Genus == "",
        Taxon,
        stringr::str_replace_all(Genus, "_", " ")
      )
    )
}

cluster_taxa_direction <- cluster_taxa_medians %>%
  group_by(Taxon) %>%
  mutate(
    Highest_cluster = Cluster_Aitchison[which.max(Median_CLR)],
    Highest_median_CLR = max(Median_CLR, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  arrange(Highest_cluster, Genus_clean, Cluster_Aitchison)

cluster_taxa_direction

cluster_taxa_summary <- cluster_taxa_medians %>%
  group_by(Taxon, Genus_clean) %>%
  slice_max(
    order_by = Median_CLR,
    n = 1,
    with_ties = FALSE
  ) %>%
  ungroup() %>%
  rename(
    Highest_cluster = Cluster_Aitchison,
    Highest_median_CLR = Median_CLR
  ) %>%
  arrange(Highest_cluster, Genus_clean)

cluster_taxa_summary

############################################################
# Part 4. Generic DAA volcano plot for clusters
############################################################
cluster_taxa_medians <- cluster_taxa_medians %>%
  dplyr::mutate(
    Genus_clean = dplyr::case_when(
      Taxon == "213ebca51a1d1b76ccd1456820b11a96548261a1" ~ "Eggerthellaceae unclassified",
      Taxon == "d7d0c7358e32837b6121cfd76b27f13f2a1edb75" ~ "Enterobacteriaceae unclassified",
      TRUE ~ Genus_clean
    )
  )

cluster_taxa_medians <- cluster_taxa_medians %>%
  mutate(
    Genus_label = case_when(
      Genus_clean == "Eggerthellaceae unclassified" ~
        "italic('Eggerthellaceae')~'unclassified'",

      Genus_clean == "Enterobacteriaceae unclassified" ~
        "italic('Enterobacteriaceae')~'unclassified'",

      TRUE ~ paste0("italic('", Genus_clean, "')")
    )
  )

cluster_taxa_plot_df <- cluster_taxa_medians %>%
  group_by(Taxon, Genus_clean) %>%
  mutate(
    Highest_cluster = Cluster_Aitchison[which.max(Median_CLR)]
  ) %>%
  ungroup() %>%
  mutate(
    Highest_cluster = factor(Highest_cluster),
    Cluster_Aitchison = factor(Cluster_Aitchison),
    Genus_clean = forcats::fct_reorder(Genus_clean, as.numeric(Highest_cluster))
  )

cluster_taxa_plot <- ggplot(
  cluster_taxa_plot_df,
  aes(
    x = Cluster_Aitchison,
    y = Genus_clean,
    fill = Median_CLR
  )
) +
  geom_tile(color = "white", linewidth = 0.6) +
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0,
    name = "Median CLR"
  ) +
  scale_y_discrete(
  labels = function(x) parse(text =
    cluster_taxa_plot_df$Genus_label[
      match(x, cluster_taxa_plot_df$Genus_clean)
    ]
  )
) +
  theme_classic(base_size = 16) +
  labs(
    x = "Aitchison cluster",
    y = NULL,
    title = "Differentially Abundant Taxa by Cluster",
    fill = "Median CLR"
  ) +
  theme(
    axis.text.y = element_text(size = 12, color = "black"),
    axis.text.x = element_text(size = 14),
    axis.title.x = element_text(size = 16),
    plot.title = element_text(size = 18, hjust = 0.5, face = "bold"),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  )
cluster_taxa_plot <- cluster_taxa_plot +
  labs(
    x = NULL,
    title = NULL
  ) +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_text(size = 18, color = "black"),
    axis.text.y = element_text(size = 13, color = "black"),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14),
    legend.key.height = unit(1.2, "cm"),
    legend.key.width = unit(0.35, "cm"),
    legend.margin = margin(0, 0, 0, 0),
    plot.margin = margin(5, 5, 5, 5)
  )
cluster_taxa_plot

```

###Save
```{r}

ggsave(
  filename = "cluster_taxa_plot.pdf",
  plot = cluster_taxa_plot,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)
```

```{r}
############################################################
# 5. Pairwise comparisons between clusters for significant taxa
############################################################

cluster_taxa_pairwise <- cluster_taxa_plot_df %>%
  dplyr::select(
    Taxon,
    Genus_clean
  ) %>%
  dplyr::distinct() %>%
  purrr::pmap_dfr(
    function(Taxon, Genus_clean) {
      
      temp <- clr_cluster_meta %>%
        dplyr::select(
          Subject.ID,
          Cluster_Aitchison,
          all_of(Taxon)
        ) %>%
        dplyr::rename(
          CLR = all_of(Taxon)
        ) %>%
        dplyr::filter(
          !is.na(Cluster_Aitchison),
          !is.na(CLR)
        )
      
      test <- pairwise.wilcox.test(
        x = temp$CLR,
        g = temp$Cluster_Aitchison,
        p.adjust.method = "fdr"
      )
      
      as.data.frame(as.table(test$p.value)) %>%
        dplyr::filter(!is.na(Freq)) %>%
        dplyr::mutate(
          Taxon = Taxon,
          Genus_clean = Genus_clean,
          Comparison = paste(Var1, "vs", Var2),
          Pairwise_FDR = Freq
        ) %>%
        dplyr::select(
          Taxon,
          Genus_clean,
          Comparison,
          Pairwise_FDR
        )
    }
  ) %>%
  dplyr::arrange(
    Genus_clean,
    Pairwise_FDR
  )

cluster_taxa_pairwise

cluster_taxa_pairwise_direction <- cluster_taxa_pairwise %>%
  tidyr::separate(
    Comparison,
    into = c("Cluster_A", "Cluster_B"),
    sep = " vs ",
    remove = FALSE
  ) %>%
  dplyr::left_join(
    cluster_taxa_medians %>%
      dplyr::rename(
        Cluster_A = Cluster_Aitchison,
        Median_CLR_A = Median_CLR
      ) %>%
      dplyr::select(Taxon, Cluster_A, Median_CLR_A),
    by = c("Taxon", "Cluster_A")
  ) %>%
  dplyr::left_join(
    cluster_taxa_medians %>%
      dplyr::rename(
        Cluster_B = Cluster_Aitchison,
        Median_CLR_B = Median_CLR
      ) %>%
      dplyr::select(Taxon, Cluster_B, Median_CLR_B),
    by = c("Taxon", "Cluster_B")
  ) %>%
  dplyr::mutate(
    Higher_cluster = dplyr::case_when(
      Median_CLR_A > Median_CLR_B ~ Cluster_A,
      Median_CLR_B > Median_CLR_A ~ Cluster_B,
      TRUE ~ "Tie"
    )
  ) %>%
  dplyr::arrange(
    Genus_clean,
    Pairwise_FDR
  )

cluster_taxa_pairwise_direction

cluster_taxa_pairwise_direction %>%
  dplyr::filter(Pairwise_FDR < 0.05)




############################################################
# 6. Plot cluster pairwise comparisons for significant taxa
############################################################

cluster_taxa_boxplot_df <- clr_cluster_meta %>%
  dplyr::select(
    Subject.ID,
    Cluster_Aitchison,
    all_of(unique(sig_zico_cluster$Taxon))
  ) %>%
  tidyr::pivot_longer(
    cols = all_of(unique(sig_zico_cluster$Taxon)),
    names_to = "Taxon",
    values_to = "CLR"
  ) %>%
  dplyr::left_join(
    sig_zico_cluster %>%
      dplyr::select(Taxon, Genus_clean),
    by = "Taxon"
  ) %>%
  dplyr::filter(
    !is.na(Cluster_Aitchison),
    !is.na(CLR)
  ) %>%
  dplyr::mutate(
    Cluster_Aitchison = factor(
      Cluster_Aitchison,
      levels = c("1", "2", "3")
    )
  )


plot_cluster_taxon <- function(taxon_name) {
  
  temp_plot <- cluster_taxa_boxplot_df %>%
    dplyr::filter(Genus_clean == taxon_name)
  
  temp_stats <- cluster_taxa_pairwise_direction %>%
    dplyr::filter(
      Genus_clean == taxon_name,
      Pairwise_FDR < 0.05
    ) %>%
    dplyr::mutate(
      group1 = Cluster_A,
      group2 = Cluster_B,
      label = dplyr::case_when(
        Pairwise_FDR < 0.001 ~ "***",
        Pairwise_FDR < 0.01 ~ "**",
        Pairwise_FDR < 0.05 ~ "*",
        TRUE ~ "ns"
      ),
      y.position = max(temp_plot$CLR, na.rm = TRUE) +
        dplyr::row_number() * 0.8
    )
  
  ggplot(
    temp_plot,
    aes(
      x = Cluster_Aitchison,
      y = CLR,
      fill = Cluster_Aitchison
    )
  ) +
    geom_boxplot(
      width = 0.65,
      outlier.shape = NA,
      color = "black",
      linewidth = 0.8
    ) +
    geom_jitter(
      width = 0.12,
      size = 3,
      color = "black",
      alpha = 0.9
    ) +
    ggpubr::stat_pvalue_manual(
      temp_stats,
      label = "label",
      y.position = "y.position",
      tip.length = 0.01,
      bracket.size = 0.8,
      size = 7
    ) +
    scale_fill_manual(
      values = c(
        "1" = "#9E2F2F",
        "2" = "#FDBB84",
        "3" = "#CB181D"
      )
    ) +
    theme_classic(base_size = 18) +
    labs(
      x = "Aitchison cluster",
      y = "CLR abundance",
      title = taxon_name
    ) +
    theme(
      plot.title = element_text(
        face = "italic",
        size = 26,
        hjust = 0.5
      ),
      axis.text.x = element_text(
        size = 18
      ),
      axis.text.y = element_text(
        size = 18
      ),
      axis.title.x = element_text(size = 22),
      axis.title.y = element_text(size = 22),
      legend.position = "none"
    )
}

dialister_plot <- plot_cluster_taxon("Dialister")
dialister_plot

dialister_df <- data.frame(
  Subject.ID = rownames(clr_mat),
  Dialister = clr_mat[, "Dialister"]
)

merged_meta_r <- merged_meta_r %>%
  left_join(dialister_df, by = "Subject.ID")

kruskal.test(
  Dialister ~ case_control_new,
  data = merged_meta_r
)

```

###Save
```{r}

ggsave(
  filename = "dialister_plot.pdf",
  plot = dialister_plot,
  path = fig_dir,
  width = 8,
  height = 5.5,
  units = "in",
  device = "pdf",
  bg = "transparent"
)
```
